---
title: "Data, Evidence, and Decision Making (with R)"
author: "Ayush Patel"
subtitle: "Session 4: Working with Data in R"
institute: "National Law School of India University, Bengaluru"
date: today
date-format: "DD MMM, YYYY"
embed-resources: true
bibliography: citations.bib
execute: 
  echo: true
format: 
  revealjs:
    fig-height: 6
    fig-width: 14
    embed-resources: true
    margin-left: 50px
    margin-right: 50px
    slide-number: c/t
    width: 1400
    height: 850
    theme: [default, custom.scss]
    footer: "Email: ayush.ap58@gmail.com"
---


# Readings

The required readings before this class:

1. Chapter 3, *Data transformation* of @wickham2023r4ds
2. Chapter 4, *Workflow: code style* of @wickham2023r4ds

## You are expected to know

1. What an object is?
2. Basic assignment in R
3. The difference between scripts and quarto documents
4. How to run code in RStudio
5. How to create vectors and data frames
6. Basic subsetting

## This Session

1. We will learn how to work with data frames
2. Use `dplyr` functions for common data manipulation tasks
3. Learn to filter, select, arrange, and create variables
4. Summarise data and count observations
5. Combine multiple steps using pipes
6. Learn about pivoting, joining, and handling missing values

# Setup

## Packages
<br><br><br>

Please Load these pckages <br>

```{r}
#| label: setup-libs
#| output: false
#| warning: false

library(tidyverse)
library(palmerpenguins)
library(flipbookr)
```

## Data - penguins

We will use the `penguins` data for most examples.

```{r}
#| label: show-penguins
#| echo: false
penguins
```

## Data - starwars

We'll also use `starwars` some examples.

```{r}
#| label: show-starwars
#| echo: false
starwars
```

## Data - mtcars

We'll also use `mtcars` some examples.

```{r}
#| label: show-mtcars
#| echo: false
mtcars
```



# Why data wrangling?

## Raw data is rarely ready

Most datasets are not immediately ready for analysis.

Usually we need to:

1. Keep only the variables we need
2. Keep only the observations we want
3. Create new variables
4. Recode old variables
5. Summarise the data

## The main verbs

The `dplyr` package gives us a few very useful functions:

1. `select()` to choose columns
2. `filter()` to choose rows
3. `arrange()` to sort rows
4. `mutate()` to create new variables
5. `summarise()` to compute summary statistics
6. `count()` to count observations
7. `group_by()` to do all of this by groups

# `filter()`

## 

![[Allison Horst](https://allisonhorst.com/)](https://raw.githubusercontent.com/allisonhorst/stats-illustrations/master/rstats-artwork/dplyr_filter.jpg)

## Filtering rows

Use `filter()` when you want only observations satisfying some condition.

```{r}
filter(starwars, height > 164)
```

## More than one condition

```{r}
filter(starwars, height > 164, mass > 80)
```

## Using logical operators

```{r}
filter(starwars, homeworld == "Naboo" | homeworld == "Tatooine")
```

## Missing values matter

If a variable has missing values, be careful.

```{r}
filter(starwars, !is.na(mass))
```

# `select()`

## Selecting variables

Use `select()` when you want only a few columns.

```{r}
select(starwars, name, films, vehicles, starships)
```

## Dropping variables

You can also remove columns by using `-`.

```{r}
select(starwars, -height, -mass)
```

## A small reminder

`select()` works on columns, not rows.

So use it when the problem is:

- "I do not need all these variables."
- "I only want these 3 or 4 columns."

## Some other valid ways 

<br><br>

```{r}
select(penguins, c(1:5))
```

## Some other valid ways 

<br><br>

```{r}
select(penguins, c(1,3,5,8))
```

## Some other valid ways 

<br><br>

```{r}
select(penguins, -c(1:5))
```

## Activity 1

   + Use the `penguins` data for this activity
   + View the `penguins` data
   + filter those observations where a penguin weighs more than 3900 gms
   + from the `penguins` data only select columns `bill_lenght_mm`, `flipper_length_mm`, `body_mass_g`
   + Get the following code to execute, it might have an error:
   
```{r eval=FALSE}
# copy this code and make it work

select(penguins,islands, c(sex,year))

```

# `arrange()`

## Sorting rows

Use `arrange()` to sort data.

```{r}
arrange(starwars, mass)
```

## Descending order

```{r}
arrange(starwars, desc(mass))
```

## Sorting by multiple variables

```{r}
#| echo: true
#| eval: false
arrange(starwars, homeworld, desc(height))
```

```{r}
#| echo: false
arrange(starwars, homeworld, desc(height)) |>
    relocate(name, height, mass, homeworld, everything())
```

# `mutate()`

## 

![[Allison Horst](https://allisonhorst.com/)](https://cdn.myportfolio.com/45214904-6a61-4e23-98d6-b140f8654a40/bd4ae264-ae51-4d18-bd60-7a058ab42fba_rw_1920.png?h=a3757d1f46f418c59e7e7946d026344e)

## Creating new variables

Use `mutate()` when you want to add a new variable.

```{r}
#| echo: true
#| eval: false
mutate(starwars,
       height_m = height / 100)
```


```{r}
#| echo: false
mutate(starwars,
       height_m = height / 100,.before = 2)
```

## Multiple variables at once

```{r}
#| echo: true
#| eval: false
mutate(
  starwars,
  height_m = height / 100,
  mass_kg = mass
)
```

```{r}
#| echo: false

mutate(
  starwars,
  height_m = height / 100,
  mass_kg = mass,
  .before = 4
)
```



## Reusing created variables

```{r}
#| echo: true
#| eval: false
mutate(
  starwars,
  height_m = height / 100,
  height_cm = height_m * 100
)
```


```{r}
#| echo: false

mutate(
  starwars,
  height_m = height / 100,
  height_cm = height_m * 100,
  .before = 4
)
```


## Why this matters

This is useful when:

1. Units need to be changed
2. Ratios or differences need to be computed
3. Variables need to be recoded before plotting or modeling

# count()

## Counting rows by group

Use `count()` when you want to know how many rows are in each group.

```{r}
count(starwars, homeworld)
```

## Sorted counts

```{r}
count(starwars, homeworld, sort = TRUE)
```

## More than one column

```{r}
count(starwars, homeworld, species)
```

## Activity 2

Use the `penguins` data to get do the following

Write code to create new columns, with appropriate names:  
  

  + that records the mass of a penguin in Kilograms 
  + column that classifies an observation as "fluffy" if it has mass more than 4500, otherwise as "not fluffy" HINT: use the `ifelse` function


# summarise()

## Getting summary values

Use `summarise()` when you want to reduce data to a few numbers.

```{r}
summarise(
  starwars,
  mean_mass = mean(mass, na.rm = TRUE),
  median_height = median(height, na.rm = TRUE)
)
```

## Common summary functions

Some common functions used inside `summarise()` are:

1. `mean()`
2. `median()`
3. `min()`
4. `max()`
5. `sd()`
6. `n()`

# group_by()

## Summaries by group

`group_by()` becomes very useful when combined with `summarise()`.

```{r}
starwars |> 
  group_by(species) |> 
  summarise(
    mean_mass = mean(mass, na.rm = TRUE),
    median_height = median(height, na.rm = TRUE),
    n = n()
  )
```

## More than one grouping variable

```{r}
starwars |> 
  group_by(homeworld, species) |> 
  summarise(
    mean_mass = mean(mass, na.rm = TRUE),
    n = n()
  )
```

## What is happening?

This code:

1. Splits the data into groups
2. Computes the summary within each group
3. Returns a new data frame


## Activity 3

Use the `penguins` data:

Write code to :  
  
  + Get mean flipper length
  + median body mass
  + the max bill lenght (HINT: use the `max` function)


# Pipes

## 


:::: {.columns}

::: {.column}

_You want to do this:_  

__f(g(h(x)))__ == __z__  
  
_f_, _g_, and _h_ are functions.  
  
_x_ is the object on which you want to use said functions  
  
_z_ is the what we get from f(g(h(x)))  


:::

::: {.column}

_Using the `%>%` or `|>`_  
  
```
x |> 
  h() |> 
    g() > 
      f()
```


:::

::::

## Why use the pipe?

1. Code readability
2. More control on intermediate steps
3. Less intermediate objects


## Why pipes?


```{r}
penguins |>
  filter(!is.na(body_mass_g)) |>
  select(species, island, body_mass_g) |>
  arrange(desc(body_mass_g))
```

## Read this slowly

Think of `|>` as:

"Take the output from the previous step and pass it into the next step."

## A more realistic example

```{r}
penguins |>
  filter(!is.na(body_mass_g), !is.na(species)) |>
  group_by(species) |>
  summarise(
    avg_mass = mean(body_mass_g, na.rm = TRUE),
    count = n()
  ) |>
  arrange(desc(avg_mass))
```

## Comparison: with vs without pipe


:::: {.columns}

::: {.column}

With `|>`:

```{r}
penguins |> 
  filter(species == "Gentoo") |> 
  mutate(
    is_fluffy = if_else(
      body_mass_g > 3500, 1, 0)
  ) |> 
  summarise(
    Number_of_fluffy_Gentoos = 
      sum(is_fluffy, na.rm = TRUE)
  )
```


:::

::: {.column}

Without `|>`:

```{r}
filter(penguins, species == "Gentoo") -> gentoo_penguins

fluffy_gentoos <- mutate(
    gentoo_penguins,
       is_fluffy = 
         if_else(body_mass_g > 3500, 1, 0)
         )

summarise(
    fluffy_gentoos,
       Number_of_fluffy_Gentoos = 
         sum(is_fluffy, na.rm = TRUE)
         )
```


:::

::::

##

```{r flip_pipe, eval=FALSE, include=FALSE,message=FALSE,warning=FALSE}

penguins %>% #BREAK
  filter(species == 
           "Gentoo") %>% #BREAK2
  mutate(
    is_fluffy = 
      if_else(body_mass_g>3500,
              1,0)
  ) %>% #BREAK3
  summarise(
    Number_of_fluffy_Gentoos = 
      sum(is_fluffy,na.rm = T),
    average_flliper_length = 
      mean(flipper_length_mm, 
           na.rm = T),
    average_bill_depth = 
      mean(bill_depth_mm,
           na.rm = T)
  )
```


`r chunk_reveal(chunk_name = "flip_pipe",platform = "quarto", title = "Peeping into the pipe")`


## Activity 4

Use the `mtcars` data.  
  
Copy the following code in your console to see the documentation about the data.  
  
```{r eval=FALSE}
?mtcars
```
  

Write code to (use ` |>` operator) :  
  
  + For cars that have manual transmission,Find the grouped average Miles per gallon for cars with various cylinders. 

# case_when()

## Grouping a variable

Use `case_when()` when you want to create a new variable based on multiple conditions.

Say you want to add a new variable according to the following requirement:

- If price is between 1 to 100 classify as `Affordable`
- If price is between 101 to 300 classify as `Not too costly`
- If price is 301 or more classify as `Will go somewhere else`

## Example

```{r}
tibble(
  food_item = c("Biryani", "Ratia", "Pasta", "Fish curry",
                "Mutton Khema", "Squid fry", "Dal chawal"),
  price = c(250, 50, 500, 600, 650, 450, 100)
) -> food_price

food_price |> 
  mutate(
    how_costly = case_when(
      price >= 1 & price <= 100 ~ "Affordable",
      price >= 101 & price <= 300 ~ "Not too costly",
      price >= 301 ~ "Will go somewhere else"
    )
  )
```

## Mutually exclusive conditions

It is best to use `case_when()` when all conditions are mutually exclusive.

```{r}
c(1, 2, 3, 4, 5, 6, 9, 8) -> numbers_obj

new_class <- case_when(
  numbers_obj > 0 ~ "Greater than 0",
  numbers_obj > 5 ~ "Greater than 5",
  numbers_obj < 100 ~ "less than 100"
)
new_class
```

## When no condition is matched

```{r}
c(numbers_obj, -1) -> numbers_obj
new_class <- case_when(
  numbers_obj > 0 ~ "Greater than 0",
  numbers_obj > 5 ~ "Greater than 5",
  numbers_obj < 100 & numbers_obj > 0 ~ 
    "less than 100 and greater than 0"
)
new_class
```

## Activity 5

Use the `penguins` data  
  
write a code to create a new column that takes the value:  
  
  + _Not fluffy_ if the body mass of the penguin is 2700-3550  
  + _Normal_ if the body mass of the penguin is 3551- 4050  
  + _Fluffy_ if the body mass of the penguin is 4751-6300  
  + Find out the frequency of the values that this new variable takes (HINT: use the `count` funciton)

# End of Session 4

# References