---
title: "Data, Evidence, and Decision Making (with R)"
author: "Ayush Patel"
subtitle: "Session 12: Walkthrough on expectations, LLN, CLT, CI, and SE"
institute: "National Law School of India University, Bengaluru"
date: today
date-format: "DD MMM, YYYY"
embed-resources: false
bibliography: citations.bib
execute: 
  echo: true
  warning: false
format: 
  revealjs:
    fig-height: 6
    fig-width: 14
    fig-retina: 1
    embed-resources: true
    margin-left: 50px
    margin-right: 50px
    slide-number: c/t
    width: 1400
    height: 850
    theme: [default, custom.scss]
    footer: "Email: ayush.ap58@gmail.com"
---

```{r}
#| label: libs
#| echo: false
#| warning: false

extrafont::loadfonts(device = "all")
library(tidyverse)
library(ggiraph)
library(hrbrthemes)

```

# Goal

> The aim for this lecture is to introduce foundational concepts in probability and statistical inference by using a simulated walk through.

# Expectation

> Over a very large number of trials, the expectation of a random process can be estimated using relative frequencies of its outcomes. 

<br>

Formally,

$E(X) = \sum_i^n {x_i P(x_i)}$

Where $X$ is a random process with $n$ outcomes, each assigned a value of $x_i$ occurring with probability $P(x_i)$

## Tossing a coin

<br>

This is a random process such that it has two outcomes: Heads and Tails. 

<br>

Imagine we had no idea of the chances of observing heads and tails in a fair coin toss.

<br>

Observing how the outcomes over a large number of tosses can allow us to learn about this.

## Tossing a coin

Let us assign 1 to heads and 0 to tails.

```{r}
#| label: show-sample-func

sample(x = c(1,0), size = 1, replace = T, prob = c(.5,.5))

sample(x = c(1,0), size = 1, replace = T, prob = c(.5,.5))

sample(x = c(1,0), size = 2, replace = T, prob = c(.5,.5))

sample(x = c(1,0), size = 10, replace = T, prob = c(.5,.5))

```

## Tossing a coin

How many heads do we get if we toss a coin 100 times?


```{r}
#| label: onehundred-tosses

sample(c(1,0), 100, T) -> hundred_tosses

hundred_tosses

sum(hundred_tosses) # total heads observed

sum(hundred_tosses)/100 # relative freq of heads

1- (sum(hundred_tosses)/100) # relative freq of tails

1*(sum(hundred_tosses)/100) + 0*(1- (sum(hundred_tosses)/100)) # an estimate of expectation of tossing a coin
```

## Tossing a coin

For a very large number of times


```{r}
#| label: tbl-toss-large-number-trials
#| output-location: slide

map(
    seq(10,10000, by = 50),
    ~ tibble(
        size = .x,
        num_heads = sample(c(1,0),.x,T) |> sum(),
        prop_head = num_heads/size
    )
) |>
    list_rbind() -> records_large_trials

records_large_trials |>
    reactable::reactable(defaultPageSize = 7)
```

## Tossing a coin


```{r}
#| label: fig-observed-lln
#| fig-cap: Proportion of heads observed across trial sizes. Shows law of large numbers.
#| output-location: slide
#| fig-dpi: 800
#| fig-height: 7 

records_large_trials |>
    ggplot(aes(size, prop_head))+
    geom_line(linetype = 2, alpha = 0.5) +
    geom_point_interactive(aes(tooltip = paste(size,"-",prop_head)),
    colour = "#880800", alpha = 0.8, size = .7) +
    labs(
        title = "Trial Size and Proportion of heads observed",
        y = "Proportion of heads",
        x = "Trial Size"
    ) +
    geom_hline(aes(yintercept = 0.5)) +
    scale_y_continuous(labels = scales::label_percent()) +
    theme_ipsum_es() -> plt_lln

girafe(ggobj = plt_lln)
```

## Should you be impressed?

<br>

If someone tells you that they observed 38% Heads form 60 coin tosses and some other person claims to have observed 54% Heads from 9860 coin tosses, which is the more impressive or rare outcome?

<br>

Is there a way to quantify this?

# Quantifying variance of observed chance

<br>

> If I toss a coin 60 times for a large number of time, and record the proportion of heads, it will give me the variability in observing heads for a given trial size (60 in this case). 

## What are the chances?


```{r}
#| label: variability-at-given-trial-size
#| fig-cap: Using the properties of a normal distribution we can quantify the chances to observe and event
#| output-location: slide
#| fig-dpi: 800
#| fig-height: 7 
map(
    c(1:5000),
    ~ tibble(
        prop_60 = sample(c(1,0), 60, T) |>
            mean(),
        prop_9860 = sample(c(1,0), 9860, T) |>
            mean()
    )
) |>
    list_rbind() -> variability_at_size

variability_at_size|>
    ggplot() +
    geom_density(aes(prop_60), colour = "#5e60d3") +
    geom_density(aes(prop_9860), colour = "#346030") +
    labs(
        title = "variability at a given trial size",
        x = "Proportion of Heads Observed"
    ) +
    scale_x_continuous(labels = scales::label_percent()) +
    theme_ipsum_es()
```

## What are the chances ? {.smaller}


```{r}
#| label: quantify-var
#| layout-ncol: 2
#| output-location: slide
#| fig-height: 8
#| fig-dpi: 800
#| column: page
#| fig-width: 7

variability_at_size|>
    ggplot() +
    geom_density(aes(prop_60),
    colour = "#5e60d3",
    fill = "#5e60d3",
    alpha = 0.5) +
    geom_vline(aes(xintercept = prop_60 |>
    quantile(.025))) +
    geom_vline(aes(xintercept = prop_60 |>
    quantile(.975))) +
    labs(
        title = "Variability at a given trial size (60)",
        subtitle = "The vertical line are the 2.5 and 97.5 percentile of the observed data",
        x = "Proportion of Heads Observed"
    ) +
    scale_x_continuous(labels = scales::label_percent()) +
    theme_ipsum_es()


variability_at_size|>
    ggplot() +
    geom_density(aes(prop_9860), 
    colour = "#346030",
    fill = "#346030",
    alpha = 0.5) +
    geom_vline(aes(xintercept = prop_9860 |>
    quantile(.025))) +
    geom_vline(aes(xintercept = prop_9860 |>
    quantile(.975))) +
    labs(
        title = "Variability at a given trial size (9860)",
        subtitle = "The vertical line are the 2.5 and 97.5 percentile of the observed data",
        x = "Proportion of Heads Observed"
    ) +
    scale_x_continuous(labels = scales::label_percent()) +
    theme_ipsum_es()

```

## What have we seen?

<br><br>

[As the number of trials increases the observed statistic approaches the expected value of the random process.]{.fragment fragment-index='1'}

<br><br>

[For a given trial size it is possible to quantify the variability due to chance in observing a statistic of interest.]{.fragment fragment-index='2'}

## The Normal Distribution

<br>

> The bell shaped curve we keep bumping into when we look at variability is called the normal distribution. It shows up so often in statistics that it is worth knowing it formally.

<br>

::: {.incremental}
- It is symmetric around its mean $\mu$
- Its spread is measured by its standard deviation $\sigma$
- Roughly 68% of values fall within 1 SD of the mean, 95% within 2 SD, and 99.7% within 3 SD
- This is called the **68-95-99.7 rule**
:::

## The Normal Distribution

$f(x) = \dfrac{1}{\sigma\sqrt{2\pi}}\, e^{-\frac{1}{2}\left(\frac{x-\mu}{\sigma}\right)^2}$

Where $\mu$ is the mean and $\sigma$ is the standard deviation of the distribution.

```{=html}
<svg viewBox="0 0 1200 480" style="width:100%;height:auto;display:block;">
  <line x1="80" y1="400" x2="1120" y2="400" stroke="#333" stroke-width="2"/>

  <path d="M100,396 C220,392 300,360 380,300 C450,250 500,150 560,100 C580,86 590,80 600,80 C610,80 620,86 640,100 C700,150 750,250 820,300 C900,360 980,392 1100,396"
        fill="none" stroke="#4C78A8" stroke-width="3"/>

  <line x1="600" y1="80" x2="600" y2="400" stroke="#666" stroke-width="1.5" stroke-dasharray="5,4"/>
  <text x="592" y="65" font-size="16" fill="#333">μ</text>

  <line x1="433" y1="150" x2="433" y2="400" stroke="#F58518" stroke-width="1.5" stroke-dasharray="5,4"/>
  <line x1="767" y1="150" x2="767" y2="400" stroke="#F58518" stroke-width="1.5" stroke-dasharray="5,4"/>
  <text x="415" y="140" font-size="14" fill="#F58518">-1σ</text>
  <text x="755" y="140" font-size="14" fill="#F58518">+1σ</text>

  <line x1="266" y1="290" x2="266" y2="400" stroke="#880800" stroke-width="1.5" stroke-dasharray="5,4"/>
  <line x1="933" y1="290" x2="933" y2="400" stroke="#880800" stroke-width="1.5" stroke-dasharray="5,4"/>
  <text x="245" y="280" font-size="14" fill="#880800">-2σ</text>
  <text x="920" y="280" font-size="14" fill="#880800">+2σ</text>

  <line x1="100" y1="396" x2="100" y2="400" stroke="#346030" stroke-width="1.5"/>
  <line x1="1100" y1="396" x2="1100" y2="400" stroke="#346030" stroke-width="1.5"/>
  <text x="70" y="420" font-size="14" fill="#346030">-3σ</text>
  <text x="1080" y="420" font-size="14" fill="#346030">+3σ</text>

  <line x1="433" y1="440" x2="767" y2="440" stroke="#F58518" stroke-width="2"/>
  <text x="555" y="460" font-size="15" fill="#F58518">68%</text>

  <line x1="266" y1="465" x2="933" y2="465" stroke="#880800" stroke-width="2"/>
  <text x="555" y="480" font-size="15" fill="#880800">95% (approx)</text>
</svg>
```

## Why This Matters for Inference {.smaller}

<br>

::: {.incremental}
- The **Central Limit Theorem** tells us the distribution of sample means is approximately normal, regardless of the shape of the population
- That means we can use the properties of the normal distribution to describe how far a sample mean is likely to fall from the true mean
- This is exactly what we did earlier using the 2.5th and 97.5th percentiles. The normal distribution gives us a formal way to do the same thing
- A **confidence interval** is built by taking the sample mean and adding and subtracting a multiple of the standard error, using the normal distribution to decide that multiple
- This is also the foundation for **hypothesis testing**, which asks how unusual an observed statistic would be if a claimed value were true
:::

# Central Limit Theorem

<br>

> Regardless of type of the population distribution, as long as the population has a finite variance, the distribution of mean of various samples of the population will closely resemble a normal distribution. 

<br>

This allows us to estimate the uncertainity of the sample estimates we observe.

## CLT

<br>

::: {.incremental}
- Take a population with **any** distribution shape (even skewed, weird, non-normal)
- Draw many random samples of size $n$ from it
- Compute the **mean** of each sample
- The distribution of those sample means becomes approximately **normal**. Even though the population isn't
:::

## Step 1: The Population {auto-animate="true"}

The population is a square full of individuals. Their values follow a skewed distribution, shown on the right.

```{=html}
<svg viewBox="0 0 1200 520" style="width:100%;height:auto;display:block;">
  <text x="60" y="45" font-size="20" font-weight="bold" fill="#333">Population</text>
  <rect x="60" y="60" width="260" height="260" fill="none" stroke="#333" stroke-width="2"/>
  <text x="60" y="345" font-size="15" fill="#666">N = 40 individuals</text>

  <text x="400" y="45" font-size="20" font-weight="bold" fill="#333">Distribution of Population Values</text>
  <line x1="400" y1="260" x2="1140" y2="260" stroke="#333" stroke-width="2"/>
  <rect x="400.0" width="57.7" y="90.0" height="170.0" fill="#4C78A8"/>
  <rect x="461.7" width="57.7" y="132.5" height="127.5" fill="#4C78A8"/>
  <rect x="523.3" width="57.7" y="185.6" height="74.4" fill="#4C78A8"/>
  <rect x="585.0" width="57.7" y="217.5" height="42.5" fill="#4C78A8"/>
  <rect x="646.7" width="57.7" y="238.8" height="21.2" fill="#4C78A8"/>
  <rect x="708.3" width="57.7" y="238.8" height="21.2" fill="#4C78A8"/>
  <rect x="893.3" width="57.7" y="249.4" height="10.6" fill="#4C78A8"/>
  <rect x="1078.3" width="57.7" y="249.4" height="10.6" fill="#4C78A8"/>
  <line x1="527.5" y1="80" x2="527.5" y2="260" stroke="#666" stroke-width="2" stroke-dasharray="6,4"/>
  <text x="533" y="95" font-size="15" fill="#666">population mean (μ)</text>

  <text x="400" y="295" font-size="20" font-weight="bold" fill="#333">Distribution of Sample Means</text>
  <line x1="400" y1="480" x2="1140" y2="480" stroke="#333" stroke-width="2"/>
  <line x1="527.5" y1="300" x2="527.5" y2="480" stroke="#666" stroke-width="2" stroke-dasharray="6,4"/>
  <text x="533" y="315" font-size="15" fill="#666">μ</text>

  <circle data-id="sq0" cx="135.0" cy="199.3" r="6" fill="#4C78A8"/>
  <circle data-id="sq1" cx="162.7" cy="211.8" r="6" fill="#4C78A8"/>
  <circle data-id="sq2" cx="216.4" cy="98.8" r="6" fill="#4C78A8"/>
  <circle data-id="sq3" cx="87.8" cy="260.9" r="6" fill="#4C78A8"/>
  <circle data-id="sq4" cx="139.5" cy="134.2" r="6" fill="#4C78A8"/>
  <circle data-id="sq5" cx="294.1" cy="183.8" r="6" fill="#4C78A8"/>
  <circle data-id="sq6" cx="260.7" cy="185.0" r="6" fill="#4C78A8"/>
  <circle data-id="sq7" cx="219.2" cy="116.6" r="6" fill="#4C78A8"/>
  <circle data-id="sq8" cx="218.3" cy="267.3" r="6" fill="#4C78A8"/>
  <circle data-id="sq9" cx="194.9" cy="240.7" r="6" fill="#4C78A8"/>
  <circle data-id="sq10" cx="226.0" cy="98.4" r="6" fill="#4C78A8"/>
  <circle data-id="sq11" cx="244.2" cy="209.1" r="6" fill="#4C78A8"/>
  <circle data-id="sq12" cx="148.3" cy="91.5" r="6" fill="#4C78A8"/>
  <circle data-id="sq13" cx="266.8" cy="184.3" r="6" fill="#4C78A8"/>
  <circle data-id="sq14" cx="236.0" cy="269.6" r="6" fill="#4C78A8"/>
  <circle data-id="sq15" cx="235.0" cy="278.4" r="6" fill="#4C78A8"/>
  <circle data-id="sq16" cx="167.9" cy="253.2" r="6" fill="#4C78A8"/>
  <circle data-id="sq17" cx="178.4" cy="281.5" r="6" fill="#4C78A8"/>
  <circle data-id="sq18" cx="269.6" cy="105.5" r="6" fill="#4C78A8"/>
  <circle data-id="sq19" cx="113.6" cy="130.6" r="6" fill="#4C78A8"/>
  <circle data-id="sq20" cx="287.8" cy="176.6" r="6" fill="#4C78A8"/>
  <circle data-id="sq21" cx="216.6" cy="148.2" r="6" fill="#4C78A8"/>
  <circle data-id="sq22" cx="191.5" cy="166.0" r="6" fill="#4C78A8"/>
  <circle data-id="sq23" cx="158.7" cy="207.9" r="6" fill="#4C78A8"/>
  <circle data-id="sq24" cx="207.7" cy="274.9" r="6" fill="#4C78A8"/>
  <circle data-id="sq25" cx="228.2" cy="280.1" r="6" fill="#4C78A8"/>
  <circle data-id="sq26" cx="264.8" cy="293.1" r="6" fill="#4C78A8"/>
  <circle data-id="sq27" cx="226.0" cy="119.3" r="6" fill="#4C78A8"/>
  <circle data-id="sq28" cx="265.7" cy="287.6" r="6" fill="#4C78A8"/>
  <circle data-id="sq29" cx="275.0" cy="204.5" r="6" fill="#4C78A8"/>
  <circle data-id="sq30" cx="234.9" cy="129.3" r="6" fill="#4C78A8"/>
  <circle data-id="sq31" cx="259.6" cy="205.4" r="6" fill="#4C78A8"/>
  <circle data-id="sq32" cx="144.8" cy="98.3" r="6" fill="#4C78A8"/>
  <circle data-id="sq33" cx="264.3" cy="292.9" r="6" fill="#4C78A8"/>
  <circle data-id="sq34" cx="103.6" cy="253.1" r="6" fill="#4C78A8"/>
  <circle data-id="sq35" cx="171.2" cy="116.7" r="6" fill="#4C78A8"/>
  <circle data-id="sq36" cx="146.7" cy="246.4" r="6" fill="#4C78A8"/>
  <circle data-id="sq37" cx="268.3" cy="94.3" r="6" fill="#4C78A8"/>
  <circle data-id="sq38" cx="214.1" cy="94.4" r="6" fill="#4C78A8"/>
  <circle data-id="sq39" cx="235.9" cy="154.5" r="6" fill="#4C78A8"/>
</svg>
```

## Step 2: Draw a Sample {auto-animate="true"}

A sample of **n = 8** individuals is drawn at random from the population square, highlighted in orange.

```{=html}
<svg viewBox="0 0 1200 520" style="width:100%;height:auto;display:block;">
  <text x="60" y="45" font-size="20" font-weight="bold" fill="#333">Population</text>
  <rect x="60" y="60" width="260" height="260" fill="none" stroke="#333" stroke-width="2"/>
  <text x="60" y="345" font-size="15" fill="#666">N = 40 individuals</text>

  <text x="400" y="45" font-size="20" font-weight="bold" fill="#333">Distribution of Population Values</text>
  <line x1="400" y1="260" x2="1140" y2="260" stroke="#333" stroke-width="2"/>
  <rect x="400.0" width="57.7" y="90.0" height="170.0" fill="#4C78A8"/>
  <rect x="461.7" width="57.7" y="132.5" height="127.5" fill="#4C78A8"/>
  <rect x="523.3" width="57.7" y="185.6" height="74.4" fill="#4C78A8"/>
  <rect x="585.0" width="57.7" y="217.5" height="42.5" fill="#4C78A8"/>
  <rect x="646.7" width="57.7" y="238.8" height="21.2" fill="#4C78A8"/>
  <rect x="708.3" width="57.7" y="238.8" height="21.2" fill="#4C78A8"/>
  <rect x="893.3" width="57.7" y="249.4" height="10.6" fill="#4C78A8"/>
  <rect x="1078.3" width="57.7" y="249.4" height="10.6" fill="#4C78A8"/>

  <line x1="527.5" y1="80" x2="527.5" y2="260" stroke="#666" stroke-width="2" stroke-dasharray="6,4"/>
  <text x="533" y="95" font-size="15" fill="#666">population mean (μ)</text>

  <text x="400" y="295" font-size="20" font-weight="bold" fill="#333">Distribution of Sample Means</text>
  <line x1="400" y1="480" x2="1140" y2="480" stroke="#333" stroke-width="2"/>
  <line x1="527.5" y1="300" x2="527.5" y2="480" stroke="#666" stroke-width="2" stroke-dasharray="6,4"/>
  <text x="533" y="315" font-size="15" fill="#666">μ</text>

  <circle data-id="sq0" cx="135.0" cy="199.3" r="6" fill="#4C78A8"/>
  <circle data-id="sq1" cx="162.7" cy="211.8" r="6" fill="#4C78A8"/>
  <circle data-id="sq2" cx="216.4" cy="98.8" r="6" fill="#4C78A8"/>
  <circle data-id="sq3" cx="87.8" cy="260.9" r="9" fill="#F58518"/>
  <circle data-id="sq4" cx="139.5" cy="134.2" r="6" fill="#4C78A8"/>
  <circle data-id="sq5" cx="294.1" cy="183.8" r="6" fill="#4C78A8"/>
  <circle data-id="sq6" cx="260.7" cy="185.0" r="6" fill="#4C78A8"/>
  <circle data-id="sq7" cx="219.2" cy="116.6" r="6" fill="#4C78A8"/>
  <circle data-id="sq8" cx="218.3" cy="267.3" r="9" fill="#F58518"/>
  <circle data-id="sq9" cx="194.9" cy="240.7" r="6" fill="#4C78A8"/>
  <circle data-id="sq10" cx="226.0" cy="98.4" r="6" fill="#4C78A8"/>
  <circle data-id="sq11" cx="244.2" cy="209.1" r="6" fill="#4C78A8"/>
  <circle data-id="sq12" cx="148.3" cy="91.5" r="6" fill="#4C78A8"/>
  <circle data-id="sq13" cx="266.8" cy="184.3" r="6" fill="#4C78A8"/>
  <circle data-id="sq14" cx="236.0" cy="269.6" r="9" fill="#F58518"/>
  <circle data-id="sq15" cx="235.0" cy="278.4" r="6" fill="#4C78A8"/>
  <circle data-id="sq16" cx="167.9" cy="253.2" r="6" fill="#4C78A8"/>
  <circle data-id="sq17" cx="178.4" cy="281.5" r="9" fill="#F58518"/>
  <circle data-id="sq18" cx="269.6" cy="105.5" r="6" fill="#4C78A8"/>
  <circle data-id="sq19" cx="113.6" cy="130.6" r="6" fill="#4C78A8"/>
  <circle data-id="sq20" cx="287.8" cy="176.6" r="6" fill="#4C78A8"/>
  <circle data-id="sq21" cx="216.6" cy="148.2" r="9" fill="#F58518"/>
  <circle data-id="sq22" cx="191.5" cy="166.0" r="6" fill="#4C78A8"/>
  <circle data-id="sq23" cx="158.7" cy="207.9" r="6" fill="#4C78A8"/>
  <circle data-id="sq24" cx="207.7" cy="274.9" r="6" fill="#4C78A8"/>
  <circle data-id="sq25" cx="228.2" cy="280.1" r="6" fill="#4C78A8"/>
  <circle data-id="sq26" cx="264.8" cy="293.1" r="9" fill="#F58518"/>
  <circle data-id="sq27" cx="226.0" cy="119.3" r="6" fill="#4C78A8"/>
  <circle data-id="sq28" cx="265.7" cy="287.6" r="6" fill="#4C78A8"/>
  <circle data-id="sq29" cx="275.0" cy="204.5" r="6" fill="#4C78A8"/>
  <circle data-id="sq30" cx="234.9" cy="129.3" r="6" fill="#4C78A8"/>
  <circle data-id="sq31" cx="259.6" cy="205.4" r="9" fill="#F58518"/>
  <circle data-id="sq32" cx="144.8" cy="98.3" r="6" fill="#4C78A8"/>
  <circle data-id="sq33" cx="264.3" cy="292.9" r="6" fill="#4C78A8"/>
  <circle data-id="sq34" cx="103.6" cy="253.1" r="6" fill="#4C78A8"/>
  <circle data-id="sq35" cx="171.2" cy="116.7" r="6" fill="#4C78A8"/>
  <circle data-id="sq36" cx="146.7" cy="246.4" r="9" fill="#F58518"/>
  <circle data-id="sq37" cx="268.3" cy="94.3" r="6" fill="#4C78A8"/>
  <circle data-id="sq38" cx="214.1" cy="94.4" r="6" fill="#4C78A8"/>
  <circle data-id="sq39" cx="235.9" cy="154.5" r="6" fill="#4C78A8"/>
</svg>
```

## The Sample Lands on the Value Axis {auto-animate="true"}

The 8 sampled individuals leave the square and take their place on the population's own value axis, at their true values.

```{=html}
<svg viewBox="0 0 1200 520" style="width:100%;height:auto;display:block;">
  <text x="60" y="45" font-size="20" font-weight="bold" fill="#333">Population</text>
  <rect x="60" y="60" width="260" height="260" fill="none" stroke="#333" stroke-width="2"/>
  <text x="60" y="345" font-size="15" fill="#666">N = 40 individuals</text>

  <text x="400" y="45" font-size="20" font-weight="bold" fill="#333">Distribution of Population Values</text>
  <line x1="400" y1="260" x2="1140" y2="260" stroke="#333" stroke-width="2"/>
  <rect x="400.0" width="57.7" y="90.0" height="170.0" fill="#4C78A8"/>
  <rect x="461.7" width="57.7" y="132.5" height="127.5" fill="#4C78A8"/>
  <rect x="523.3" width="57.7" y="185.6" height="74.4" fill="#4C78A8"/>
  <rect x="585.0" width="57.7" y="217.5" height="42.5" fill="#4C78A8"/>
  <rect x="646.7" width="57.7" y="238.8" height="21.2" fill="#4C78A8"/>
  <rect x="708.3" width="57.7" y="238.8" height="21.2" fill="#4C78A8"/>
  <rect x="893.3" width="57.7" y="249.4" height="10.6" fill="#4C78A8"/>
  <rect x="1078.3" width="57.7" y="249.4" height="10.6" fill="#4C78A8"/>
  <line x1="527.5" y1="80" x2="527.5" y2="260" stroke="#666" stroke-width="2" stroke-dasharray="6,4"/>
  <text x="533" y="95" font-size="15" fill="#666">population mean (μ)</text>

  <line x1="511.5" y1="215" x2="511.5" y2="260" stroke="#F58518" stroke-width="2" stroke-dasharray="4,3"/>
  <text x="365" y="210" font-size="15" fill="#F58518">this sample's mean x̄ ≈ 511</text>

  <text x="400" y="295" font-size="20" font-weight="bold" fill="#333">Distribution of Sample Means</text>
  <line x1="400" y1="480" x2="1140" y2="480" stroke="#333" stroke-width="2"/>
  <line x1="527.5" y1="300" x2="527.5" y2="480" stroke="#666" stroke-width="2" stroke-dasharray="6,4"/>
  <text x="533" y="315" font-size="15" fill="#666">μ</text>

  <circle data-id="sq0" cx="135.0" cy="199.3" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq1" cx="162.7" cy="211.8" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq2" cx="216.4" cy="98.8" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq3" cx="410.9" cy="250" r="8" fill="#F58518"/>
  <circle data-id="sq4" cx="139.5" cy="134.2" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq5" cx="294.1" cy="183.8" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq6" cx="260.7" cy="185.0" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq7" cx="219.2" cy="116.6" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq8" cx="422.8" cy="250" r="8" fill="#F58518"/>
  <circle data-id="sq9" cx="194.9" cy="240.7" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq10" cx="226.0" cy="98.4" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq11" cx="244.2" cy="209.1" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq12" cx="148.3" cy="91.5" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq13" cx="266.8" cy="184.3" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq14" cx="445.9" cy="250" r="8" fill="#F58518"/>
  <circle data-id="sq15" cx="235.0" cy="278.4" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq16" cx="167.9" cy="253.2" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq17" cx="468.6" cy="250" r="8" fill="#F58518"/>
  <circle data-id="sq18" cx="269.6" cy="105.5" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq19" cx="113.6" cy="130.6" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq20" cx="287.8" cy="176.6" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq21" cx="500.5" cy="250" r="8" fill="#F58518"/>
  <circle data-id="sq22" cx="191.5" cy="166.0" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq23" cx="158.7" cy="207.9" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq24" cx="207.7" cy="274.9" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq25" cx="228.2" cy="280.1" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq26" cx="539.6" cy="250" r="8" fill="#F58518"/>
  <circle data-id="sq27" cx="226.0" cy="119.3" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq28" cx="265.7" cy="287.6" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq29" cx="275.0" cy="204.5" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq30" cx="234.9" cy="129.3" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq31" cx="585.2" cy="250" r="8" fill="#F58518"/>
  <circle data-id="sq32" cx="144.8" cy="98.3" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq33" cx="264.3" cy="292.9" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq34" cx="103.6" cy="253.1" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq35" cx="171.2" cy="116.7" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq36" cx="718.8" cy="250" r="8" fill="#F58518"/>
  <circle data-id="sq37" cx="268.3" cy="94.3" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq38" cx="214.1" cy="94.4" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq39" cx="235.9" cy="154.5" r="6" fill="#4C78A8" opacity="0.15"/>
</svg>
```

## The Sample Mean {auto-animate="true"}

Those 8 values are averaged. The mean drops down and becomes **one point** in a new chart: the distribution of sample means.

```{=html}
<svg viewBox="0 0 1200 520" style="width:100%;height:auto;display:block;">
  <text x="60" y="45" font-size="20" font-weight="bold" fill="#333">Population</text>
  <rect x="60" y="60" width="260" height="260" fill="none" stroke="#333" stroke-width="2"/>
  <text x="60" y="345" font-size="15" fill="#666">N = 40 individuals</text>

  <text x="400" y="45" font-size="20" font-weight="bold" fill="#333">Distribution of Population Values</text>
  <line x1="400" y1="260" x2="1140" y2="260" stroke="#333" stroke-width="2"/>
  <rect x="400.0" width="57.7" y="90.0" height="170.0" fill="#4C78A8"/>
  <rect x="461.7" width="57.7" y="132.5" height="127.5" fill="#4C78A8"/>
  <rect x="523.3" width="57.7" y="185.6" height="74.4" fill="#4C78A8"/>
  <rect x="585.0" width="57.7" y="217.5" height="42.5" fill="#4C78A8"/>
  <rect x="646.7" width="57.7" y="238.8" height="21.2" fill="#4C78A8"/>
  <rect x="708.3" width="57.7" y="238.8" height="21.2" fill="#4C78A8"/>
  <rect x="893.3" width="57.7" y="249.4" height="10.6" fill="#4C78A8"/>
  <rect x="1078.3" width="57.7" y="249.4" height="10.6" fill="#4C78A8"/>
  <line x1="527.5" y1="80" x2="527.5" y2="260" stroke="#666" stroke-width="2" stroke-dasharray="6,4"/>
  <text x="533" y="95" font-size="15" fill="#666">population mean (μ)</text>

  <text x="400" y="295" font-size="20" font-weight="bold" fill="#333">Distribution of Sample Means</text>
  <line x1="400" y1="480" x2="1140" y2="480" stroke="#333" stroke-width="2"/>
  <line x1="527.5" y1="300" x2="527.5" y2="480" stroke="#666" stroke-width="2" stroke-dasharray="6,4"/>
  <text x="533" y="315" font-size="15" fill="#666">μ</text>
  <text x="420" y="465" font-size="15" fill="crimson">x̄ ≈ 511</text>

  <circle data-id="sq0" cx="135.0" cy="199.3" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq1" cx="162.7" cy="211.8" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq2" cx="216.4" cy="98.8" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq3" cx="511" cy="452" r="4" fill="crimson"/>
  <circle data-id="sq4" cx="139.5" cy="134.2" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq5" cx="294.1" cy="183.8" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq6" cx="260.7" cy="185.0" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq7" cx="219.2" cy="116.6" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq8" cx="511" cy="452" r="4" fill="crimson"/>
  <circle data-id="sq9" cx="194.9" cy="240.7" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq10" cx="226.0" cy="98.4" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq11" cx="244.2" cy="209.1" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq12" cx="148.3" cy="91.5" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq13" cx="266.8" cy="184.3" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq14" cx="511" cy="452" r="4" fill="crimson"/>
  <circle data-id="sq15" cx="235.0" cy="278.4" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq16" cx="167.9" cy="253.2" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq17" cx="511" cy="452" r="4" fill="crimson"/>
  <circle data-id="sq18" cx="269.6" cy="105.5" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq19" cx="113.6" cy="130.6" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq20" cx="287.8" cy="176.6" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq21" cx="511" cy="452" r="4" fill="crimson"/>
  <circle data-id="sq22" cx="191.5" cy="166.0" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq23" cx="158.7" cy="207.9" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq24" cx="207.7" cy="274.9" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq25" cx="228.2" cy="280.1" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq26" cx="511" cy="452" r="4" fill="crimson"/>
  <circle data-id="sq27" cx="226.0" cy="119.3" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq28" cx="265.7" cy="287.6" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq29" cx="275.0" cy="204.5" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq30" cx="234.9" cy="129.3" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq31" cx="511" cy="452" r="4" fill="crimson"/>
  <circle data-id="sq32" cx="144.8" cy="98.3" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq33" cx="264.3" cy="292.9" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq34" cx="103.6" cy="253.1" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq35" cx="171.2" cy="116.7" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq36" cx="511" cy="452" r="4" fill="crimson"/>
  <circle data-id="sq37" cx="268.3" cy="94.3" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq38" cx="214.1" cy="94.4" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq39" cx="235.9" cy="154.5" r="6" fill="#4C78A8" opacity="0.15"/>
</svg>
```

## Step 3: Repeat the Process

Do this hundreds of times. Click through to watch the sample means pile up into a bell shape.

```{=html}
<svg viewBox="0 0 1200 520" style="width:100%;height:auto;display:block;">
  <text x="400" y="45" font-size="20" font-weight="bold" fill="#333">Distribution of Sample Means</text>
  <line x1="400" y1="480" x2="1140" y2="480" stroke="#333" stroke-width="2"/>
  <line x1="527.5" y1="80" x2="527.5" y2="480" stroke="#666" stroke-width="2" stroke-dasharray="6,4"/>
  <text x="533" y="95" font-size="15" fill="#666">population mean (μ)</text>

  <!-- the one mean we already drew, at x = 511 -->
  <circle cx="511" cy="472" r="7" fill="crimson"/>

  <g class="fragment fade-in">
    <circle cx="511" cy="458" r="7" fill="crimson"/><circle cx="511" cy="444" r="7" fill="crimson"/>
    <circle cx="511" cy="430" r="7" fill="crimson"/><circle cx="511" cy="416" r="7" fill="crimson"/>
    <circle cx="511" cy="402" r="7" fill="crimson"/><circle cx="511" cy="388" r="7" fill="crimson"/>
    <circle cx="511" cy="374" r="7" fill="crimson"/><circle cx="511" cy="360" r="7" fill="crimson"/>
    <circle cx="511" cy="346" r="7" fill="crimson"/><circle cx="511" cy="332" r="7" fill="crimson"/>
    <circle cx="511" cy="318" r="7" fill="crimson"/>
  </g>

  <g class="fragment fade-in">
    <circle cx="481" cy="472" r="7" fill="crimson"/><circle cx="481" cy="458" r="7" fill="crimson"/>
    <circle cx="481" cy="444" r="7" fill="crimson"/><circle cx="481" cy="430" r="7" fill="crimson"/>
    <circle cx="481" cy="416" r="7" fill="crimson"/><circle cx="481" cy="402" r="7" fill="crimson"/>
    <circle cx="481" cy="388" r="7" fill="crimson"/><circle cx="481" cy="374" r="7" fill="crimson"/>
    <circle cx="481" cy="360" r="7" fill="crimson"/><circle cx="481" cy="346" r="7" fill="crimson"/>
    <circle cx="541" cy="472" r="7" fill="crimson"/><circle cx="541" cy="458" r="7" fill="crimson"/>
    <circle cx="541" cy="444" r="7" fill="crimson"/><circle cx="541" cy="430" r="7" fill="crimson"/>
    <circle cx="541" cy="416" r="7" fill="crimson"/><circle cx="541" cy="402" r="7" fill="crimson"/>
    <circle cx="541" cy="388" r="7" fill="crimson"/><circle cx="541" cy="374" r="7" fill="crimson"/>
    <circle cx="541" cy="360" r="7" fill="crimson"/><circle cx="541" cy="346" r="7" fill="crimson"/>
  </g>

  <g class="fragment fade-in">
    <circle cx="451" cy="472" r="7" fill="crimson"/><circle cx="451" cy="458" r="7" fill="crimson"/>
    <circle cx="451" cy="444" r="7" fill="crimson"/><circle cx="451" cy="430" r="7" fill="crimson"/>
    <circle cx="451" cy="416" r="7" fill="crimson"/><circle cx="451" cy="402" r="7" fill="crimson"/>
    <circle cx="451" cy="388" r="7" fill="crimson"/>
    <circle cx="571" cy="472" r="7" fill="crimson"/><circle cx="571" cy="458" r="7" fill="crimson"/>
    <circle cx="571" cy="444" r="7" fill="crimson"/><circle cx="571" cy="430" r="7" fill="crimson"/>
    <circle cx="571" cy="416" r="7" fill="crimson"/><circle cx="571" cy="402" r="7" fill="crimson"/>
    <circle cx="571" cy="388" r="7" fill="crimson"/>
  </g>

  <g class="fragment fade-in">
    <circle cx="421" cy="472" r="7" fill="crimson"/><circle cx="421" cy="458" r="7" fill="crimson"/>
    <circle cx="421" cy="444" r="7" fill="crimson"/><circle cx="421" cy="430" r="7" fill="crimson"/>
    <circle cx="601" cy="472" r="7" fill="crimson"/><circle cx="601" cy="458" r="7" fill="crimson"/>
    <circle cx="601" cy="444" r="7" fill="crimson"/><circle cx="601" cy="430" r="7" fill="crimson"/>
  </g>

  <g class="fragment fade-in">
    <circle cx="391" cy="472" r="7" fill="crimson"/><circle cx="391" cy="458" r="7" fill="crimson"/>
    <circle cx="631" cy="472" r="7" fill="crimson"/><circle cx="631" cy="458" r="7" fill="crimson"/>
  </g>

  <g class="fragment fade-in">
    <circle cx="361" cy="472" r="7" fill="crimson"/>
    <circle cx="661" cy="472" r="7" fill="crimson"/>
  </g>

  <g class="fragment fade-in">
    <path d="M 340 478 C 420 478, 450 300, 511 300 S 602 478, 682 478"
          fill="none" stroke="#333" stroke-width="3"/>
    <text x="720" y="340" font-size="22" fill="#333" font-weight="bold">Approximately Normal</text>
  </g>
</svg>
```

## Takeaway

::: {.incremental}
- The **population** can look like anything. Here, it's skewed
- Each **individual sample mean** bounces around near the true mean
- The **distribution of sample means**, across many samples, is approximately $N\left(\mu, \dfrac{\sigma}{\sqrt{n}}\right)$
- Bigger $n$ → tighter, more normal-looking sampling distribution
:::


# Lets take it for a spin

<br>

> The total assets declared by general election candidates can be found on the ADR website. All the nearly 8000 candidates is actually the populaiton. We can decide on a sample size of 80 and see how well CLT works.

## Test CLT: an example

<br><br>

:::{.incremental}

1. Load data of all ~ 8000 general election candidates
2. draw a sample of size `n` and calcualte sample mean of assets.
3. repeat step 2 for a large number of times (K)
4. Plot density chart of all K sample means.
5. Does it look normal?
:::

## Test CLT: an example

**Load data**


```{r}
#| label: load-data-election
#| tbl-cap: Data from ADR on asset declared by general election candidates
#| output-location: slide
#| column: body
#| warning: false

read_csv("Lok-sabha-2024-candidates-details.csv") -> data_ele

data_ele |>
    select(sno, candidate, party, criminal_case) -> data_ele



data_ele |>
    reactable::reactable(defaultPageSize = 3) 
```

## Test CLT: an example


```{r}
#| label: show-samp

data_ele$party |>
    sample(size = 80, replace = F)

data_ele$party |>
    sample(size = 80, replace = F)
```

## Test CLT: an example


```{r}
#| label: tbl-create-samp-dist
#| tbl-cap: 10000 resamples of size 300
#| output-location: slide

map(
    c(1:10000),
    ~tibble(
        sample = .x, 
        prop_bjp = (data_ele$party == "BJP")|>
            sample(size = 300, replace = F) |>
            mean(na.rm = T),
        mean_criminal_case = data_ele$criminal_case |>
            sample(size = 300, replace = F) |>
            mean(na.rm = T)

    )
) |>
    list_rbind() -> dist_samples

dist_samples|>
    reactable::reactable(defaultPageSize = 4)
```

## Test CLT: an example


```{r}
#| label: fig-samp-dist-of-mean
#| fig-cap: Sampling distribution comes out to be close to normal, this is CLT in work
#| output-location: slide
#| layout-ncol: 2
#| fig-height: 6
#| fig-dpi: 800
#| column: page
#| fig-width: 7

dist_samples |>
    ggplot(aes(prop_bjp)) +
    geom_density(fill = "steelblue", alpha = 0.7) +
    labs(
        title = "Disribution of 1000 sample means",
        x = "Proportion of candidates that contested on a BJP ticket"
    ) +
        theme_ipsum_es()

dist_samples |>
    ggplot(aes(prop_bjp)) +
    geom_density(fill = "steelblue", alpha = 0.7) +
    labs(
        title = "Disribution of 1000 sample means",
        x = "Average Criminal Cases"
    ) +
        theme_ipsum_es()
```

# Check Your Understanding

<br>

Quick questions to test the ideas from today. Think first, then click to reveal.

## Question 1

<br>

If you toss a fair coin 20 times, is the proportion of heads you observe likely to be exactly 0.5?

<br>

[Not necessarily. For small trial sizes the observed proportion can differ a fair bit from the true expectation. It is only as the number of trials grows large that the proportion reliably gets close to 0.5. This is the law of large numbers.]{.fragment}

## Question 2 {.smaller}

<br>

Two people report their coin toss results. Person A observed 65% heads from 40 tosses. Person B observed 53% heads from 5000 tosses. Whose result should surprise you more, and why?

<br>

[Person B's result is actually more surprising, even though 53% looks closer to 50% than Person A's 65%. At small trial sizes the variability in observed proportions is much higher, so a proportion like 65% from just 40 tosses is still fairly ordinary by chance. At large trial sizes the variability shrinks a great deal, so even a small deviation like 53% from 5000 tosses is many standard errors away from 0.5, and far less likely to happen by chance alone. What matters is not the raw distance from 0.5, but that distance measured in standard errors, which shrink as the sample size grows.]{.fragment}

## Question 3

<br>

A friend says, "The Central Limit Theorem tells us that any large enough dataset will look normally distributed." What is wrong with this statement?

<br>

[This confuses the population distribution with the sampling distribution. The CLT is about the distribution of sample means computed across repeated samples, not about individual data points. A single large dataset drawn from a skewed population stays skewed. It is the distribution of means of many such samples that becomes approximately normal.]{.fragment}

## Question 4

<br>

Why does increasing the sample size make the sampling distribution of the mean narrower?

<br>

[The standard error of the mean is $\sigma/\sqrt{n}$. As $n$ grows, the denominator grows, so the standard error shrinks. Sample means from larger samples vary less from one sample to the next, so the sampling distribution tightens around the true mean.]{.fragment}

## Question 5

<br>

Roughly what proportion of sample means would you expect to fall within 2 standard errors of the true mean, if the sampling distribution is approximately normal?

<br>

[About 95%. This follows from the 68-95-99.7 rule for the normal distribution, applied to the sampling distribution of the mean rather than to individual observations.]{.fragment}

## Question 6 {.smaller}

Suppose a country has 300 million people, and only 300 of them are actual terrorists. That is a base rate of 1 in a million.

The algorithm correctly flags 99% of actual terrorists.

It wrongly flags 1% of innocent people as well.

a) How many actual terrorists get flagged?

b) How many innocent people get flagged?

c) If a person is flagged, what is the probability they are actually a terrorist?

# R Building Blocks: Loops and map()

<br>

> Before the task ahead, a quick refresher on three ways to repeat an operation in R. We have already been using `map()` throughout today's session, here is what it is doing under the hood.

## for loops

A `for` loop repeats an operation a fixed number of times, one step at a time.

```{r}
#| label: for-loop-example

sample_sizes <- c(10, 50, 100, 500)
mean_heads <- vector("double", length(sample_sizes))

for (i in seq_along(sample_sizes)) {
    mean_heads[i] <- sample(c(1,0), sample_sizes[i], replace = TRUE) |> mean()
}

mean_heads
```

Each pass through the loop fills in one slot of `mean_heads`, using the sample size at that same position in `sample_sizes`.

## while loops 

A `while` loop repeats an operation until a condition is no longer true. Useful when you do not know in advance how many repeats you will need.

```{r}
#| label: while-loop-example

set.seed(42)
tosses <- 0
outcome <- 0

while (outcome != 1) {
    outcome <- sample(c(1,0), 1, replace = TRUE)
    tosses <- tosses + 1
}

tosses
```

This keeps tossing a coin, one at a time, until it finally lands heads, and reports how many tosses that took.

## map(): the tidyverse way

`map()` applies the same operation to every element of a list or vector, and hands back the results together. This is what we have been using all session, for example to build `records_large_trials` earlier.

```{r}
#| label: map-example

map_dbl(
    c(10, 50, 100, 500),
    ~ sample(c(1,0), .x, replace = TRUE) |> mean()
)
```

Same result as the `for` loop version, in fewer lines, and without having to set up an empty vector first.

## Which One Should You Use? {.smaller}

<br>

::: {.incremental}
- **`for`**: full control over each step, useful when a step depends on what happened in a previous step
- **`while`**: repeat until a condition is met, useful when the number of repeats is not known in advance
- **`map()`**: the cleanest choice when applying the same operation across a list of inputs and collecting the results, this is what we used throughout today's walkthrough
- All three are doing the same underlying thing: repetition. The choice is about what makes your code easiest to read and to get right
:::

# Apply What You've Learned

<br>

> An authentic task, similar in spirit to what you will do in your midterm and final projects.

## Authentic Task

<br>

You are given `bail-amounts-data.csv`, a dataset of bail amounts set by lower courts over the last five years, covering just over 4,000 cases. A newspaper claims that "the average bail amount in 2025 was 20% higher than in 2024."

## Authentic Task: What To Do {.smaller}

<br>

::: {.incremental}
- Filter `data_bail` to 2024 and 2025 using the `year` column, and work with `bail_amount_inr`
- Draw repeated random samples of a fixed size from each year's cases and compute the sample mean bail amount each time, using either a loop or `map()`
- Plot the two sampling distributions of the mean and describe their shape and spread
- Use the standard error to construct an approximate range within which you would expect the true average to fall, for each year
- State, in plain language, whether the data supports the newspaper's claim, and how confident you are in that conclusion
- Note any way the sampling process itself, for example which cases get recorded or omitted, might bias your answer
:::

# End of Session