---
title: "Data, Evidence, and Decision Making (with R)"
author: "Ayush Patel"
subtitle: "Session 6: Data Visualization with R"
institute: "National Law School of India University, Bengaluru"
date: today
date-format: "DD MMM, YYYY"
embed-resources: true
bibliography: citations.bib
execute: 
  echo: true
  warning: false
format: 
  revealjs:
    fig-height: 6
    fig-width: 14
    fig-dpi: 800
    embed-resources: true
    margin-left: 50px
    margin-right: 50px
    slide-number: c/t
    width: 1400
    height: 850
    theme: [default, custom.scss]
    footer: "Email: ayush.ap58@gmail.com"
---

# Readings

1. Hadley Wickham, Mine Çetinkaya-Rundel, andGarrett Grolemund. R for Data Science (2nd ed.). Chapter 1: Data Visualization. @wickham2023r4ds

2. Healy, Kieran. Data Visualization: A Practical Introduction. Chapter 4: Show the right numbers. Sections 4.1 to 4.3. @healy2019socviz

# Goal

1. Learn to make charts with {ggplot2}
2. Understand the grammar of graphics
3. To make charts for exploratory, descriptive, and storytelling

# Load these libraries

```{r}
library(tidyverse)
library(gapminder)
```

# The dataset

<br>

> We will be using `gapminder` and `msleep` data. Use the command `?gapminder` or `?msleep` to learn more about these datasets.

## The dataset (gapminder) {.smaller}

```{r}
#| label: tbl-show-gapminder
#| tbl-cap: "Gapminder Data"
#| echo: false

gapminder |>
    reactable::reactable()
```

## The dataset (msleep) {.smaller}

```{r}
#| label: tbl-show-msleep
#| tbl-cap: "Mammals Sleep Data"
#| echo: false

msleep |>
    reactable::reactable(defaultPageSize = 5, 
    compact = T, 
    width = "auto",
    height = "auto",
    theme = reactable::reactableTheme(
        style = list(
            fontSize = "smaller" # Or use specific values like "12px"
            )
        )
    )
```


## Components of a ggplot - canvas {auto-animate=true}

```{r}
#| echo: true
#| output-location: column
#| fig-width: 8
#| fig-height: 6
#| code-line-numbers: "2"

# Data to use for creating plots
ggplot(data = msleep) 

```

## Components of a ggplot - link vars to aesthetics {auto-animate=true}

```{r}
#| echo: true
#| output-location: column
#| fig-width: 8
#| fig-height: 6
#| code-line-numbers: "4,5,6"

# Data to use for creating plots
ggplot(data = msleep, 
    # link variables in data to axes
       aes(x = bodywt, 
           y = sleep_total) 
       ) 

```

## Components of a ggplot - Choose geometry {auto-animate=true}

```{r}
#| echo: true
#| output-location: column
#| fig-width: 8
#| fig-height: 6
#| code-line-numbers: "6,8"

# Data to use for creating plots
ggplot(data = msleep, 
    # link variables in data to axes
       aes(x = bodywt, 
           y = sleep_total) 
       )+ # plus sign'+' adds a layer
    # A geometry or type of chart to plot
  geom_point() 


```

## Components of a ggplot - the details {auto-animate=true}

```{r}
#| echo: true
#| output-location: column
#| fig-width: 8
#| fig-height: 6
#| code-line-numbers: "9-18"

# Data to use for creating plots
ggplot(data = msleep, 
    # link variables in data to axes
       aes(x = bodywt, 
           y = sleep_total) 
       )+ # plus sign'+' adds a layer
    # A geometry or type of chart to plot
  geom_point() + # A '+' sign for every layer
  # add or edit various labels with this function
  labs( 
    # title for plot 
    title = "Are body weight and sleep time related?", 
    # x axis title
    x = "Body weight (Kg)", 
    # y axis title,
    y = "Total sleep time (hours)" 
  )+
  theme_minimal() # cosmetics


```

## Summary - commponents of a ggplot {auto-animate=true}

Create a blank canvas by using `ggplot()`

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "1"

ggplot(data = df) # df is any data frame 
```

## Summary - commponents of a ggplot {auto-animate=true}

Link variables (columns) from df to plot aesthetics using `aes()`

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "2-5"

ggplot(data = df, # df is any data frame 
       aes(
         x = var1, # var1 is name of a column in df
         y = var2  # var1 is name of a column in df
         )
       ) 
```

## Summary - commponents of a ggplot {auto-animate=true}

Choose a geometry using `geom_*()`

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "7"

ggplot(data = df, # df is any data frame 
       aes(
         x = var1, # var1 is name of a column in df
         y = var2  # var1 is name of a column in df
         )
       ) +
  geom_point() # geom layer, there are many geoms to chhoose from
```

## Summary - commponents of a ggplot {auto-animate=true}

Add label details and cosmetics

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "8-15"

ggplot(data = df, # df is any data frame 
       aes(
         x = var1, # var1 is name of a column in df
         y = var2  # var1 is name of a column in df
         )
       ) +
  geom_point() +# geom layer, there are many geoms to choose from
  labs(
    title = "Plot title",
    subtitle = "Plot Sub-title",
    x = "x-axis title",
    y = "y-axis title",
    caption = "plot caption"
  )+
  theme_bw() # there are many themes to choose from, you can make your own as well
```

## A quick coding exercise - 1 {{< fa solid laptop-code >}}

Write a code to create a plot with following details:

  1. Use the `msleep` data. (execute `?msleep` to see details of the data set)
  2. A plot between variables `vore` and `sleep_total`
  3. Link `vore` to x-axis and `sleep_total` to y-axis
  4. Use geometry boxplot (hint: `geom_boxplot`)
  5. Add appropriate titles and labels to the plot
  6. Use the `theme_*()` options, choose any

## Solution - 1 {{< fa solid check >}} {.nostrech}

```{r}
#| echo: true
#| output-location: slide
#| fig-align: center
#| fig-width: 12
#| fig-height: 6

ggplot(data = msleep,
       aes(
         x = vore,
         y = sleep_total
         )
       )+
  geom_boxplot(alpha = 0.5)+
  labs(
    title = "Sleep patters of animals by vore",
    subtitle = "Broadly speaking, animals that eat insects sleep the most",
    x = "Type of food",
    y = "Total hours of sleep in a day",
    caption = "data: V. M. Savage and G. B. West. A quantitative, theoretical framework for understanding mammalian sleep "
  )+
  theme_light()

```

## Exploring Geometries 

. . .

<br>
There are several geometries to choose from.
<br>
<br>

. . .

Which geometry is appropriate for a given plot?
<br>
<br>

. . .

We will discuss a few essential geometries and their use cases

## Situation - One Discrete variable

[Plot frequencies of conservation status of animals in `msleep` data.]{style="font-size:20px"} Use case - `geom_bar()`

```{r}
#| code-line-numbers: "3|5"
#| echo: true
#| fig-align: center
#| output-location: column-fragment
#| fig-width: 8
#| fig-height: 6

ggplot(
  data = msleep,
  aes(x = conservation)
)+
  geom_bar()+
  theme_bw()

```

## Situation - One Continuous variable

[Plot sleep time of animals in `msleep` data.]{style="font-size:20px"} Use case - `geom_density()`

```{r}
#| code-line-numbers: "3|5"
#| echo: true
#| fig-align: center
#| output-location: column-fragment
#| fig-width: 8
#| fig-height: 6

ggplot(
  data = msleep,
  aes(x = sleep_total)
)+
  geom_density()+
  theme_bw()

```

## Situation - One Continuous variable 

[Plot sleep time of animals in `msleep` data.]{style="font-size:20px"} Use case - `geom_histogram()`

```{r}
#| code-line-numbers: "3|5"
#| echo: true
#| fig-align: center
#| output-location: column-fragment
#| fig-width: 8
#| fig-height: 6

ggplot(
  data = msleep,
  aes(x = sleep_total)
)+
  geom_histogram()+
  theme_bw()

```

## Situation - Two Continuous variable

[Plot sleep time of animals and rem sleep time in `msleep` data.]{style="font-size:20px"} Use case - `geom_point()`

```{r}
#| code-line-numbers: "3-4|6"
#| echo: true
#| fig-align: center
#| output-location: column-fragment
#| fig-width: 8
#| fig-height: 6

ggplot(
  data = msleep,
  aes(x = sleep_total,
      y = sleep_rem)
)+
  geom_point()+
  theme_bw()

```

## Situation - Two Continuous variable

[Plot sleep time of animals and rem sleep time in `msleep` data.]{style="font-size:20px"} Use case - `geom_smooth()`

```{r}
#| code-line-numbers: "3-4|6"
#| echo: true
#| fig-align: center
#| output-location: column-fragment
#| fig-width: 8
#| fig-height: 6

ggplot(
  data = msleep,
  aes(x = sleep_total,
      y = sleep_rem)
)+
  geom_smooth()+
  theme_bw()

```


## Situation - 1 Continuous, 1 Discrete

[Plot sleep time of animals and conservation status in `msleep` data.]{style="font-size:20px"} Use case - `geom_col()`

```{r}
#| code-line-numbers: "1-5|7-8|10"
#| echo: true
#| fig-align: center
#| output-location: column-fragment
#| fig-width: 8
#| fig-height: 6


msleep|>
  group_by(conservation)|>
  summarise(
    avg_sleep_time = mean(sleep_total,na.rm = T)
  )|>
  ggplot(
  aes(x = conservation,
      y = avg_sleep_time)
  )+
  geom_col()+
  theme_bw()

```

## Situation - 1 Continuous, 1 Discrete

[Plot sleep time of animals and conservation status in `msleep` data.]{style="font-size:20px"} Use case - `geom_boxplot()`

```{r}
#| code-line-numbers: "3-4|6"
#| echo: true
#| fig-align: center
#| output-location: column-fragment
#| fig-width: 8
#| fig-height: 6

ggplot(
  data = msleep,
  aes(x = conservation,
      y = sleep_total)
)+
  geom_boxplot()+
  theme_bw()

```


## Summary - Geometries

```{r}
#| echo: false

tibble(
  `Situation` = c("1 Discrete Variable",
                  "1 Continious Varialbe",
                  "2 Continious Variable",
                  "1 Continiouos,1 Discrete"),
  `Geom Options` = c("geom_bar()",
                     "geom_density(), geom_histogram()",
                     "geom_point(), geom_smooth()",
                     "geom_col(), geom_boxplot()")
)|>
    reactable::reactable()

```

## A quick coding detour - 2 {{< fa solid laptop-code >}}

Use the data `gapminder` and create a boxplot using variables `continent` and `lifeExp`

## Solution - 2 {{< fa solid check >}}

```{r}
#| echo: true
#| output-location: slide
#| fig-align: center
#| fig-width: 12
#| fig-height: 6

ggplot(data = gapminder,
       aes(x = continent,
           y = lifeExp
           )
       )+
  geom_boxplot()+
  labs(
    title = "Distribution of life expectancy across continents",
    x = "Continent",
    y = "Life expectancy in years"
  )+
  theme_bw()

```


## A quick coding detour - 3 {{< fa solid laptop-code >}}

Use the data `gapminder` and create a point/scatter chart using variables `gdpPercap` and `lifeExp`

## Solution - 3 {{< fa solid check >}}

```{r}
#| echo: true
#| output-location: slide
#| fig-align: center
#| fig-width: 12
#| fig-height: 6

ggplot(data = gapminder,
       aes(x = gdpPercap,
           y = lifeExp
           )
       )+
  geom_point()+
  labs(
    title = "GDP and LIfe expectancy",
    x = "GDP per capita in USD",
    y = "Life expectancy in years"
  )+
  theme_bw()

```

## Aesthetics for geometries


 `size` | `colour` | `fill` | `alpha`
 
 . . .
  
_Why ?_

  - aesthetics can be linked with variables in data - using `aes()`
  - values can be assigned to aesthetics directly
  - makes the plots look more appealing
  - helps add more nuance to the plots

. . . 

> Some of the following slides are meant to show what is possible, does not mean one should do it all at once


## Aesthetics use case - size {auto-animate=true}

original chart

```{r}
#| echo: true
#| output-location: column-fragment
#| fig-align: center
#| fig-width: 8
#| fig-height: 6
#| code-line-numbers: "6"

ggplot(data = gapminder,
       aes(x = gdpPercap,
           y = lifeExp
           )
       )+
  geom_point()+
  labs(
    title = "GDP and LIfe expectancy",
    x = "GDP per capita in USD",
    y = "Life expectancy in years"
  )+
  theme_bw()

```

## Aesthetics use case - size {auto-animate=true}

Supply values directly to `size`

```{r}
#| echo: true
#| output-location: column-fragment
#| fig-align: center
#| fig-width: 8
#| fig-height: 6
#| code-line-numbers: "6"

ggplot(data = gapminder,
       aes(x = gdpPercap,
           y = lifeExp
           )
       )+
  geom_point(size = 3)+
  labs(
    title = "GDP and LIfe expectancy",
    x = "GDP per capita in USD",
    y = "Life expectancy in years"
  )+
  theme_bw()

```

## Aesthetics use case - size {auto-animate=true}

link `size` to a variable in data

```{r}
#| echo: true
#| output-location: column-fragment
#| fig-align: center
#| fig-width: 8
#| fig-height: 6
#| code-line-numbers: "6"

ggplot(data = gapminder,
       aes(x = gdpPercap,
           y = lifeExp
           )
       )+
  geom_point(aes(size = pop))+
  labs(
    title = "GDP and LIfe expectancy",
    x = "GDP per capita in USD",
    y = "Life expectancy in years"
  )+
  theme_bw()

```

## Aesthetics use case - alpha {auto-animate=true}

original chart

```{r}
#| echo: true
#| output-location: column-fragment
#| fig-align: center
#| fig-width: 8
#| fig-height: 6
#| code-line-numbers: "6"

ggplot(data = gapminder,
       aes(x = gdpPercap,
           y = lifeExp
           )
       )+
  geom_point()+
  labs(
    title = "GDP and LIfe expectancy",
    x = "GDP per capita in USD",
    y = "Life expectancy in years"
  )+
  theme_bw()

```

## Aesthetics use case - alpha {auto-animate=true}

Supply values directly to `alpha`

```{r}
#| echo: true
#| output-location: column-fragment
#| fig-align: center
#| fig-width: 8
#| fig-height: 6
#| code-line-numbers: "6"

ggplot(data = gapminder,
       aes(x = gdpPercap,
           y = lifeExp
           )
       )+
  geom_point(alpha = 0.5)+
  labs(
    title = "GDP and LIfe expectancy",
    x = "GDP per capita in USD",
    y = "Life expectancy in years"
  )+
  theme_bw()

```

## Aesthetics use case - colour {auto-animate=true}

original chart

```{r}
#| echo: true
#| output-location: column-fragment
#| fig-align: center
#| fig-width: 8
#| fig-height: 6
#| code-line-numbers: "6"

ggplot(data = gapminder,
       aes(x = gdpPercap,
           y = lifeExp
           )
       )+
  geom_point()+
  labs(
    title = "GDP and LIfe expectancy",
    x = "GDP per capita in USD",
    y = "Life expectancy in years"
  )+
  theme_bw()

```

## Aesthetics use case - colour {auto-animate=true}

Supply values directly to `colour`

```{r}
#| echo: true
#| output-location: column-fragment
#| fig-align: center
#| fig-width: 8
#| fig-height: 6
#| code-line-numbers: "6"

ggplot(data = gapminder,
       aes(x = gdpPercap,
           y = lifeExp
           )
       )+
  geom_point(colour = "red")+
  labs(
    title = "GDP and LIfe expectancy",
    x = "GDP per capita in USD",
    y = "Life expectancy in years"
  )+
  theme_bw()

```

## Aesthetics use case - colour {auto-animate=true}

link `colour` to a variable in data

```{r}
#| echo: true
#| output-location: column-fragment
#| fig-align: center
#| fig-width: 8
#| fig-height: 6
#| code-line-numbers: "6"

ggplot(data = gapminder,
       aes(x = gdpPercap,
           y = lifeExp
           )
       )+
  geom_point(aes(colour = continent))+
  labs(
    title = "GDP and LIfe expectancy",
    x = "GDP per capita in USD",
    y = "Life expectancy in years"
  )+
  theme_bw()

```

## Aesthetics - `fill` vs `colour`

::::{.columns}

:::{.column width="50%"}
```{r}
#| echo: true
#| code-line-numbers: "6"

ggplot(
  data = msleep,
  aes(x = conservation,
      y = sleep_total)
)+
  geom_boxplot(fill = "red")+
  theme_bw()
```

:::

:::{.column width="50%"}
```{r}
#| echo: true
#| code-line-numbers: "6"

ggplot(
  data = msleep,
  aes(x = conservation,
      y = sleep_total)
)+
  geom_boxplot(colour = "red")+
  theme_bw()
```

:::

::::

## Aesthetics - `fill` vs `colour`

::::{.columns}

:::{.column width="50%"}
```{r}
#| echo: true
#| code-line-numbers: "6"

ggplot(
  data = msleep,
  aes(x = conservation,
      y = sleep_total)
)+
  geom_boxplot(aes(fill = vore))+
  theme_bw()
```

:::

:::{.column width="50%"}
```{r}
#| echo: true
#| code-line-numbers: "6"

ggplot(
  data = msleep,
  aes(x = conservation,
      y = sleep_total)
)+
  geom_boxplot(aes(colour = vore))+
  theme_bw()
```

:::

::::

## You can have it all

**Doesnt mean you should do it**

```{r}
#| echo: true
#| output-location: slide
#| fig-align: center
#| fig-width: 14
#| fig-height: 6
#| code-line-numbers: "6-9"


ggplot(
  data = msleep,
  aes(x = conservation,
      y = sleep_total)
)+
  geom_jitter(aes(colour = vore, 
                  size = bodywt),
              alpha = 0.7)+
  geom_boxplot(aes(colour = vore))+
  theme_bw()


```


## A quick coding detour - 4 {{< fa solid laptop-code >}}

[Write a code that will generate the following chart. Use `gapminder` data.]{style="font-size:30px"}

```{r}
#| echo: false
#| code-line-numbers: "2,3,4,5"
#| fig-align: center
#| fig-width: 14
#| fig-height: 6

ggplot(gapminder,
       aes(x = lifeExp))+
  geom_density(aes(fill = continent),
               alpha = 0.5)+
  scale_fill_manual(values = continent_colors)+
  labs(
    title = "Distribution of Life Expectancy",
    y = "Density",
    x = "Life Expectancy in years"
  )+
  theme_bw()

```


## Solution - 4 {{< fa solid check >}}

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "2,3,4,5"


ggplot(gapminder,
       aes(x = lifeExp))+
  geom_density(aes(fill = continent),
               alpha = 0.5)+
  scale_fill_manual(values = continent_colors)+
  labs(
    title = "Distribution of Life Expectancy",
    y = "Density",
    x = "Life Expectancy in years"
  )+
  theme_bw()

```



## Stacked bar charts {auto-animate=true}

```{r}
#| echo: true
#| output-location: column
#| fig-width: 8
#| fig-height: 6
#| code-line-numbers: "2"

ggplot(msleep)+
  geom_bar(aes(conservation))+
  theme_bw()

```

## Stacked bar charts {auto-animate=true}

```{r}
#| echo: true
#| output-location: column
#| fig-width: 8
#| fig-height: 6
#| code-line-numbers: "2"

ggplot(msleep)+
  geom_bar(aes(conservation, fill = vore))+
  theme_bw()

```

## Stacked bar charts {auto-animate=true}

```{r}
#| echo: true
#| output-location: column
#| fig-width: 8
#| fig-height: 6
#| code-line-numbers: "2,3"

ggplot(msleep)+
  geom_bar(aes(conservation, fill = vore),
           position = "dodge")+
  theme_bw()

```

## Stacked bar charts {auto-animate=true}

```{r}
#| echo: true
#| output-location: column
#| fig-width: 8
#| fig-height: 6
#| code-line-numbers: "2,3"

ggplot(msleep)+
  geom_bar(aes(conservation, fill = vore),
           position = "fill")+
  theme_bw()

```

# Learn More

It is left to you to explore these slides to learn other details of plotting using ggplot2

1. [Deck1](https://ayushbipinpatel.github.io/GIPE-Business-Analytics/Lecture3_DataVisualization.html#/title-slide)
2. [Deck2](https://ayushbipinpatel.github.io/GIPE-Business-Analytics/Lecture9_AdvancedDataVisualizations.html#/title-slide)
3. [Deck3](https://ayushbipinpatel.github.io/GIPE-Business-Analytics/Lecture10_PresentingDataVisualizations.html#/title-slide) 


# Reference
