---
title: "Data, Evidence, and Decision Making (with R)"
author: "Ayush Patel"
subtitle: "Session 12: Walkthrough on expectations, LLN, CLT, CI, and SE"
institute: "National Law School of India University, Bengaluru"
date: today
date-format: "DD MMM, YYYY"
embed-resources: false
bibliography: citations.bib
execute: 
  echo: true
  warning: false
format: 
  revealjs:
    fig-height: 6
    fig-width: 14
    fig-retina: 1
    embed-resources: true
    margin-left: 50px
    margin-right: 50px
    slide-number: c/t
    width: 1400
    height: 850
    theme: [default, custom.scss]
    footer: "Email: ayush.ap58@gmail.com"
---

```{r}
#| label: libs
#| echo: false
#| warning: false

extrafont::loadfonts(device = "all")
library(tidyverse)
library(ggiraph)
library(hrbrthemes)

```

# Goal

> The aim for this lecture is to introduce foundational concepts in probability and statistical inference by using a simulated walk through.

# Expectation

> Over a very large number of trials, the expectation of a random process can be estimated using relative frequencies of its outcomes. 

<br>

Formally,

$E(X) = \sum_i^n {x_i P(x_i)}$

Where $X$ is a random process with $n$ outcomes, each assigned a value of $x_i$ occurring with probability $P(x_i)$

## Tossing a coin

<br>

This is a random process such that it has two outcomes: Heads and Tails. 

<br>

Imagine we had no idea of the chances of observing heads and tails in a fair coin toss.

<br>

Observing how the outcomes over a large number of tosses can allow us to learn about this.

## Tossing a coin

Let us assign 1 to heads and 0 to tails.

```{r}
#| label: show-sample-func

sample(x = c(1,0), size = 1, replace = T, prob = c(.5,.5))

sample(x = c(1,0), size = 1, replace = T, prob = c(.5,.5))

sample(x = c(1,0), size = 2, replace = T, prob = c(.5,.5))

sample(x = c(1,0), size = 10, replace = T, prob = c(.5,.5))

```

## Tossing a coin

How many heads do we get if we toss a coin 100 times?


```{r}
#| label: onehundred-tosses

sample(c(1,0), 100, T) -> hundred_tosses

hundred_tosses

sum(hundred_tosses) # total heads observed

sum(hundred_tosses)/100 # relative freq of heads

1- (sum(hundred_tosses)/100) # relative freq of tails

1*(sum(hundred_tosses)/100) + 0*(1- (sum(hundred_tosses)/100)) # an estimate of expectation of tossing a coin
```

## Tossing a coin

For a very large number of times


```{r}
#| label: tbl-toss-large-number-trials
#| output-location: slide

map(
    seq(10,10000, by = 50),
    ~ tibble(
        size = .x,
        num_heads = sample(c(1,0),.x,T) |> sum(),
        prop_head = num_heads/size
    )
) |>
    list_rbind() -> records_large_trials

records_large_trials |>
    reactable::reactable(defaultPageSize = 7)
```

## Tossing a coin


```{r}
#| label: fig-observed-lln
#| fig-cap: Proportion of heads observed across trial sizes. Shows law of large numbers.
#| output-location: slide
#| fig-dpi: 800
#| fig-height: 7 

records_large_trials |>
    ggplot(aes(size, prop_head))+
    geom_line(linetype = 2, alpha = 0.5) +
    geom_point_interactive(aes(tooltip = paste(size,"-",prop_head)),
    colour = "#880800", alpha = 0.8, size = .7) +
    labs(
        title = "Trial Size and Proportion of heads observed",
        y = "Proportion of heads",
        x = "Trial Size"
    ) +
    geom_hline(aes(yintercept = 0.5)) +
    scale_y_continuous(labels = scales::label_percent()) +
    theme_ipsum_es() -> plt_lln

girafe(ggobj = plt_lln)
```

## Should you be impressed?

<br>

If someone tells you that they observed 38% Heads form 60 coin tosses and some other person claims to have observed 54% Heads from 9860 coin tosses, which is the more impressive or rare outcome?

<br>

Is there a way to quantify this?

# Quantifying variance of observed chance

<br>

> If I toss a coin 60 times for a large number of time, and record the proportion of heads, it will give me the variability in observing heads for a given trial size (60 in this case). 

## What are the chances?


```{r}
#| label: variability-at-given-trial-size
#| fig-cap: Using the properties of a normal distribution we can quantify the chances to observe and event
#| output-location: slide
#| fig-dpi: 800
#| fig-height: 7 
map(
    c(1:5000),
    ~ tibble(
        prop_60 = sample(c(1,0), 60, T) |>
            mean(),
        prop_9860 = sample(c(1,0), 9860, T) |>
            mean()
    )
) |>
    list_rbind() -> variability_at_size

variability_at_size|>
    ggplot() +
    geom_density(aes(prop_60), colour = "#5e60d3") +
    geom_density(aes(prop_9860), colour = "#346030") +
    labs(
        title = "variability at a given trial size",
        x = "Proportion of Heads Observed"
    ) +
    scale_x_continuous(labels = scales::label_percent()) +
    theme_ipsum_es()
```

## What are the chances ? {.smaller}


```{r}
#| label: quantify-var
#| layout-ncol: 2
#| output-location: slide
#| fig-height: 8
#| fig-dpi: 800
#| column: page
#| fig-width: 7

variability_at_size|>
    ggplot() +
    geom_density(aes(prop_60),
    colour = "#5e60d3",
    fill = "#5e60d3",
    alpha = 0.5) +
    geom_vline(aes(xintercept = prop_60 |>
    quantile(.025))) +
    geom_vline(aes(xintercept = prop_60 |>
    quantile(.975))) +
    labs(
        title = "Variability at a given trial size (60)",
        subtitle = "The vertical line are the 2.5 and 97.5 percentile of the observed data",
        x = "Proportion of Heads Observed"
    ) +
    scale_x_continuous(labels = scales::label_percent()) +
    theme_ipsum_es()


variability_at_size|>
    ggplot() +
    geom_density(aes(prop_9860), 
    colour = "#346030",
    fill = "#346030",
    alpha = 0.5) +
    geom_vline(aes(xintercept = prop_9860 |>
    quantile(.025))) +
    geom_vline(aes(xintercept = prop_9860 |>
    quantile(.975))) +
    labs(
        title = "Variability at a given trial size (9860)",
        subtitle = "The vertical line are the 2.5 and 97.5 percentile of the observed data",
        x = "Proportion of Heads Observed"
    ) +
    scale_x_continuous(labels = scales::label_percent()) +
    theme_ipsum_es()

```

## What have we seen?

<br><br>

[As the number of trials increases the observed statistic approaches the expected value of the random process.]{.fragment fragment-index='1'}

<br><br>

[For a given trial size it is possible to quantify the variability due to chance in observing a statistic of interest.]{.fragment fragment-index='2'}

# Central Limit Theorem

<br>

> Regardless of type of the population distribution, as long as the population has a finite variance, the distribution of mean of various samples of the population will closely resemble a normal distribution. 

<br>

This allows us to estimate the uncertainity of the sample estimates we observe.

## CLT

<br>

::: {.incremental}
- Take a population with **any** distribution shape (even skewed, weird, non-normal)
- Draw many random samples of size $n$ from it
- Compute the **mean** of each sample
- The distribution of those sample means becomes approximately **normal**. Even though the population isn't
:::

## Step 1: The Population {auto-animate="true"}

The population is a square full of individuals. Their values follow a skewed distribution, shown on the right.

```{=html}
<svg viewBox="0 0 1200 520" style="width:100%;height:auto;display:block;">
  <text x="60" y="45" font-size="20" font-weight="bold" fill="#333">Population</text>
  <rect x="60" y="60" width="260" height="260" fill="none" stroke="#333" stroke-width="2"/>
  <text x="60" y="345" font-size="15" fill="#666">N = 40 individuals</text>

  <text x="400" y="45" font-size="20" font-weight="bold" fill="#333">Distribution of Population Values</text>
  <line x1="400" y1="260" x2="1140" y2="260" stroke="#333" stroke-width="2"/>
  <rect x="400.0" width="57.7" y="90.0" height="170.0" fill="#4C78A8"/>
  <rect x="461.7" width="57.7" y="132.5" height="127.5" fill="#4C78A8"/>
  <rect x="523.3" width="57.7" y="185.6" height="74.4" fill="#4C78A8"/>
  <rect x="585.0" width="57.7" y="217.5" height="42.5" fill="#4C78A8"/>
  <rect x="646.7" width="57.7" y="238.8" height="21.2" fill="#4C78A8"/>
  <rect x="708.3" width="57.7" y="238.8" height="21.2" fill="#4C78A8"/>
  <rect x="893.3" width="57.7" y="249.4" height="10.6" fill="#4C78A8"/>
  <rect x="1078.3" width="57.7" y="249.4" height="10.6" fill="#4C78A8"/>

  <text x="400" y="295" font-size="20" font-weight="bold" fill="#333">Distribution of Sample Means</text>
  <line x1="400" y1="480" x2="1140" y2="480" stroke="#333" stroke-width="2"/>

  <circle data-id="sq0" cx="135.0" cy="199.3" r="6" fill="#4C78A8"/>
  <circle data-id="sq1" cx="162.7" cy="211.8" r="6" fill="#4C78A8"/>
  <circle data-id="sq2" cx="216.4" cy="98.8" r="6" fill="#4C78A8"/>
  <circle data-id="sq3" cx="87.8" cy="260.9" r="6" fill="#4C78A8"/>
  <circle data-id="sq4" cx="139.5" cy="134.2" r="6" fill="#4C78A8"/>
  <circle data-id="sq5" cx="294.1" cy="183.8" r="6" fill="#4C78A8"/>
  <circle data-id="sq6" cx="260.7" cy="185.0" r="6" fill="#4C78A8"/>
  <circle data-id="sq7" cx="219.2" cy="116.6" r="6" fill="#4C78A8"/>
  <circle data-id="sq8" cx="218.3" cy="267.3" r="6" fill="#4C78A8"/>
  <circle data-id="sq9" cx="194.9" cy="240.7" r="6" fill="#4C78A8"/>
  <circle data-id="sq10" cx="226.0" cy="98.4" r="6" fill="#4C78A8"/>
  <circle data-id="sq11" cx="244.2" cy="209.1" r="6" fill="#4C78A8"/>
  <circle data-id="sq12" cx="148.3" cy="91.5" r="6" fill="#4C78A8"/>
  <circle data-id="sq13" cx="266.8" cy="184.3" r="6" fill="#4C78A8"/>
  <circle data-id="sq14" cx="236.0" cy="269.6" r="6" fill="#4C78A8"/>
  <circle data-id="sq15" cx="235.0" cy="278.4" r="6" fill="#4C78A8"/>
  <circle data-id="sq16" cx="167.9" cy="253.2" r="6" fill="#4C78A8"/>
  <circle data-id="sq17" cx="178.4" cy="281.5" r="6" fill="#4C78A8"/>
  <circle data-id="sq18" cx="269.6" cy="105.5" r="6" fill="#4C78A8"/>
  <circle data-id="sq19" cx="113.6" cy="130.6" r="6" fill="#4C78A8"/>
  <circle data-id="sq20" cx="287.8" cy="176.6" r="6" fill="#4C78A8"/>
  <circle data-id="sq21" cx="216.6" cy="148.2" r="6" fill="#4C78A8"/>
  <circle data-id="sq22" cx="191.5" cy="166.0" r="6" fill="#4C78A8"/>
  <circle data-id="sq23" cx="158.7" cy="207.9" r="6" fill="#4C78A8"/>
  <circle data-id="sq24" cx="207.7" cy="274.9" r="6" fill="#4C78A8"/>
  <circle data-id="sq25" cx="228.2" cy="280.1" r="6" fill="#4C78A8"/>
  <circle data-id="sq26" cx="264.8" cy="293.1" r="6" fill="#4C78A8"/>
  <circle data-id="sq27" cx="226.0" cy="119.3" r="6" fill="#4C78A8"/>
  <circle data-id="sq28" cx="265.7" cy="287.6" r="6" fill="#4C78A8"/>
  <circle data-id="sq29" cx="275.0" cy="204.5" r="6" fill="#4C78A8"/>
  <circle data-id="sq30" cx="234.9" cy="129.3" r="6" fill="#4C78A8"/>
  <circle data-id="sq31" cx="259.6" cy="205.4" r="6" fill="#4C78A8"/>
  <circle data-id="sq32" cx="144.8" cy="98.3" r="6" fill="#4C78A8"/>
  <circle data-id="sq33" cx="264.3" cy="292.9" r="6" fill="#4C78A8"/>
  <circle data-id="sq34" cx="103.6" cy="253.1" r="6" fill="#4C78A8"/>
  <circle data-id="sq35" cx="171.2" cy="116.7" r="6" fill="#4C78A8"/>
  <circle data-id="sq36" cx="146.7" cy="246.4" r="6" fill="#4C78A8"/>
  <circle data-id="sq37" cx="268.3" cy="94.3" r="6" fill="#4C78A8"/>
  <circle data-id="sq38" cx="214.1" cy="94.4" r="6" fill="#4C78A8"/>
  <circle data-id="sq39" cx="235.9" cy="154.5" r="6" fill="#4C78A8"/>
</svg>
```

## Step 2: Draw a Sample {auto-animate="true"}

A sample of **n = 8** individuals is drawn at random from the population square, highlighted in orange.

```{=html}
<svg viewBox="0 0 1200 520" style="width:100%;height:auto;display:block;">
  <text x="60" y="45" font-size="20" font-weight="bold" fill="#333">Population</text>
  <rect x="60" y="60" width="260" height="260" fill="none" stroke="#333" stroke-width="2"/>
  <text x="60" y="345" font-size="15" fill="#666">N = 40 individuals</text>

  <text x="400" y="45" font-size="20" font-weight="bold" fill="#333">Distribution of Population Values</text>
  <line x1="400" y1="260" x2="1140" y2="260" stroke="#333" stroke-width="2"/>
  <rect x="400.0" width="57.7" y="90.0" height="170.0" fill="#4C78A8"/>
  <rect x="461.7" width="57.7" y="132.5" height="127.5" fill="#4C78A8"/>
  <rect x="523.3" width="57.7" y="185.6" height="74.4" fill="#4C78A8"/>
  <rect x="585.0" width="57.7" y="217.5" height="42.5" fill="#4C78A8"/>
  <rect x="646.7" width="57.7" y="238.8" height="21.2" fill="#4C78A8"/>
  <rect x="708.3" width="57.7" y="238.8" height="21.2" fill="#4C78A8"/>
  <rect x="893.3" width="57.7" y="249.4" height="10.6" fill="#4C78A8"/>
  <rect x="1078.3" width="57.7" y="249.4" height="10.6" fill="#4C78A8"/>

  <text x="400" y="295" font-size="20" font-weight="bold" fill="#333">Distribution of Sample Means</text>
  <line x1="400" y1="480" x2="1140" y2="480" stroke="#333" stroke-width="2"/>

  <circle data-id="sq0" cx="135.0" cy="199.3" r="6" fill="#4C78A8"/>
  <circle data-id="sq1" cx="162.7" cy="211.8" r="6" fill="#4C78A8"/>
  <circle data-id="sq2" cx="216.4" cy="98.8" r="6" fill="#4C78A8"/>
  <circle data-id="sq3" cx="87.8" cy="260.9" r="9" fill="#F58518"/>
  <circle data-id="sq4" cx="139.5" cy="134.2" r="6" fill="#4C78A8"/>
  <circle data-id="sq5" cx="294.1" cy="183.8" r="6" fill="#4C78A8"/>
  <circle data-id="sq6" cx="260.7" cy="185.0" r="6" fill="#4C78A8"/>
  <circle data-id="sq7" cx="219.2" cy="116.6" r="6" fill="#4C78A8"/>
  <circle data-id="sq8" cx="218.3" cy="267.3" r="9" fill="#F58518"/>
  <circle data-id="sq9" cx="194.9" cy="240.7" r="6" fill="#4C78A8"/>
  <circle data-id="sq10" cx="226.0" cy="98.4" r="6" fill="#4C78A8"/>
  <circle data-id="sq11" cx="244.2" cy="209.1" r="6" fill="#4C78A8"/>
  <circle data-id="sq12" cx="148.3" cy="91.5" r="6" fill="#4C78A8"/>
  <circle data-id="sq13" cx="266.8" cy="184.3" r="6" fill="#4C78A8"/>
  <circle data-id="sq14" cx="236.0" cy="269.6" r="9" fill="#F58518"/>
  <circle data-id="sq15" cx="235.0" cy="278.4" r="6" fill="#4C78A8"/>
  <circle data-id="sq16" cx="167.9" cy="253.2" r="6" fill="#4C78A8"/>
  <circle data-id="sq17" cx="178.4" cy="281.5" r="9" fill="#F58518"/>
  <circle data-id="sq18" cx="269.6" cy="105.5" r="6" fill="#4C78A8"/>
  <circle data-id="sq19" cx="113.6" cy="130.6" r="6" fill="#4C78A8"/>
  <circle data-id="sq20" cx="287.8" cy="176.6" r="6" fill="#4C78A8"/>
  <circle data-id="sq21" cx="216.6" cy="148.2" r="9" fill="#F58518"/>
  <circle data-id="sq22" cx="191.5" cy="166.0" r="6" fill="#4C78A8"/>
  <circle data-id="sq23" cx="158.7" cy="207.9" r="6" fill="#4C78A8"/>
  <circle data-id="sq24" cx="207.7" cy="274.9" r="6" fill="#4C78A8"/>
  <circle data-id="sq25" cx="228.2" cy="280.1" r="6" fill="#4C78A8"/>
  <circle data-id="sq26" cx="264.8" cy="293.1" r="9" fill="#F58518"/>
  <circle data-id="sq27" cx="226.0" cy="119.3" r="6" fill="#4C78A8"/>
  <circle data-id="sq28" cx="265.7" cy="287.6" r="6" fill="#4C78A8"/>
  <circle data-id="sq29" cx="275.0" cy="204.5" r="6" fill="#4C78A8"/>
  <circle data-id="sq30" cx="234.9" cy="129.3" r="6" fill="#4C78A8"/>
  <circle data-id="sq31" cx="259.6" cy="205.4" r="9" fill="#F58518"/>
  <circle data-id="sq32" cx="144.8" cy="98.3" r="6" fill="#4C78A8"/>
  <circle data-id="sq33" cx="264.3" cy="292.9" r="6" fill="#4C78A8"/>
  <circle data-id="sq34" cx="103.6" cy="253.1" r="6" fill="#4C78A8"/>
  <circle data-id="sq35" cx="171.2" cy="116.7" r="6" fill="#4C78A8"/>
  <circle data-id="sq36" cx="146.7" cy="246.4" r="9" fill="#F58518"/>
  <circle data-id="sq37" cx="268.3" cy="94.3" r="6" fill="#4C78A8"/>
  <circle data-id="sq38" cx="214.1" cy="94.4" r="6" fill="#4C78A8"/>
  <circle data-id="sq39" cx="235.9" cy="154.5" r="6" fill="#4C78A8"/>
</svg>
```

## The Sample Mean {auto-animate="true"}

Those 8 values are averaged. The mean becomes **one point** in a new chart: the distribution of sample means.

```{=html}
<svg viewBox="0 0 1200 520" style="width:100%;height:auto;display:block;">
  <text x="60" y="45" font-size="20" font-weight="bold" fill="#333">Population</text>
  <rect x="60" y="60" width="260" height="260" fill="none" stroke="#333" stroke-width="2"/>
  <text x="60" y="345" font-size="15" fill="#666">N = 40 individuals</text>

  <text x="400" y="45" font-size="20" font-weight="bold" fill="#333">Distribution of Population Values</text>
  <line x1="400" y1="260" x2="1140" y2="260" stroke="#333" stroke-width="2"/>
  <rect x="400.0" width="57.7" y="90.0" height="170.0" fill="#4C78A8"/>
  <rect x="461.7" width="57.7" y="132.5" height="127.5" fill="#4C78A8"/>
  <rect x="523.3" width="57.7" y="185.6" height="74.4" fill="#4C78A8"/>
  <rect x="585.0" width="57.7" y="217.5" height="42.5" fill="#4C78A8"/>
  <rect x="646.7" width="57.7" y="238.8" height="21.2" fill="#4C78A8"/>
  <rect x="708.3" width="57.7" y="238.8" height="21.2" fill="#4C78A8"/>
  <rect x="893.3" width="57.7" y="249.4" height="10.6" fill="#4C78A8"/>
  <rect x="1078.3" width="57.7" y="249.4" height="10.6" fill="#4C78A8"/>

  <text x="400" y="295" font-size="20" font-weight="bold" fill="#333">Distribution of Sample Means</text>
  <line x1="400" y1="480" x2="1140" y2="480" stroke="#333" stroke-width="2"/>

  <circle data-id="sq0" cx="135.0" cy="199.3" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq1" cx="162.7" cy="211.8" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq2" cx="216.4" cy="98.8" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq3" cx="511" cy="452" r="4" fill="crimson"/>
  <circle data-id="sq4" cx="139.5" cy="134.2" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq5" cx="294.1" cy="183.8" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq6" cx="260.7" cy="185.0" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq7" cx="219.2" cy="116.6" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq8" cx="511" cy="452" r="4" fill="crimson"/>
  <circle data-id="sq9" cx="194.9" cy="240.7" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq10" cx="226.0" cy="98.4" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq11" cx="244.2" cy="209.1" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq12" cx="148.3" cy="91.5" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq13" cx="266.8" cy="184.3" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq14" cx="511" cy="452" r="4" fill="crimson"/>
  <circle data-id="sq15" cx="235.0" cy="278.4" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq16" cx="167.9" cy="253.2" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq17" cx="511" cy="452" r="4" fill="crimson"/>
  <circle data-id="sq18" cx="269.6" cy="105.5" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq19" cx="113.6" cy="130.6" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq20" cx="287.8" cy="176.6" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq21" cx="511" cy="452" r="4" fill="crimson"/>
  <circle data-id="sq22" cx="191.5" cy="166.0" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq23" cx="158.7" cy="207.9" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq24" cx="207.7" cy="274.9" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq25" cx="228.2" cy="280.1" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq26" cx="511" cy="452" r="4" fill="crimson"/>
  <circle data-id="sq27" cx="226.0" cy="119.3" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq28" cx="265.7" cy="287.6" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq29" cx="275.0" cy="204.5" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq30" cx="234.9" cy="129.3" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq31" cx="511" cy="452" r="4" fill="crimson"/>
  <circle data-id="sq32" cx="144.8" cy="98.3" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq33" cx="264.3" cy="292.9" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq34" cx="103.6" cy="253.1" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq35" cx="171.2" cy="116.7" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq36" cx="511" cy="452" r="4" fill="crimson"/>
  <circle data-id="sq37" cx="268.3" cy="94.3" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq38" cx="214.1" cy="94.4" r="6" fill="#4C78A8" opacity="0.15"/>
  <circle data-id="sq39" cx="235.9" cy="154.5" r="6" fill="#4C78A8" opacity="0.15"/>
</svg>
```

## Step 3: Repeat the Process

Do this hundreds of times. Click through to watch the sample means pile up into a bell shape.

```{=html}
<svg viewBox="0 0 1200 520" style="width:100%;height:auto;display:block;">
  <text x="400" y="45" font-size="20" font-weight="bold" fill="#333">Distribution of Sample Means</text>
  <line x1="400" y1="480" x2="1140" y2="480" stroke="#333" stroke-width="2"/>

  <!-- the one mean we already drew, at x = 511 -->
  <circle cx="511" cy="472" r="7" fill="crimson"/>

  <g class="fragment fade-in">
    <circle cx="511" cy="458" r="7" fill="crimson"/><circle cx="511" cy="444" r="7" fill="crimson"/>
    <circle cx="511" cy="430" r="7" fill="crimson"/><circle cx="511" cy="416" r="7" fill="crimson"/>
    <circle cx="511" cy="402" r="7" fill="crimson"/><circle cx="511" cy="388" r="7" fill="crimson"/>
    <circle cx="511" cy="374" r="7" fill="crimson"/><circle cx="511" cy="360" r="7" fill="crimson"/>
    <circle cx="511" cy="346" r="7" fill="crimson"/><circle cx="511" cy="332" r="7" fill="crimson"/>
    <circle cx="511" cy="318" r="7" fill="crimson"/>
  </g>

  <g class="fragment fade-in">
    <circle cx="481" cy="472" r="7" fill="crimson"/><circle cx="481" cy="458" r="7" fill="crimson"/>
    <circle cx="481" cy="444" r="7" fill="crimson"/><circle cx="481" cy="430" r="7" fill="crimson"/>
    <circle cx="481" cy="416" r="7" fill="crimson"/><circle cx="481" cy="402" r="7" fill="crimson"/>
    <circle cx="481" cy="388" r="7" fill="crimson"/><circle cx="481" cy="374" r="7" fill="crimson"/>
    <circle cx="481" cy="360" r="7" fill="crimson"/><circle cx="481" cy="346" r="7" fill="crimson"/>
    <circle cx="541" cy="472" r="7" fill="crimson"/><circle cx="541" cy="458" r="7" fill="crimson"/>
    <circle cx="541" cy="444" r="7" fill="crimson"/><circle cx="541" cy="430" r="7" fill="crimson"/>
    <circle cx="541" cy="416" r="7" fill="crimson"/><circle cx="541" cy="402" r="7" fill="crimson"/>
    <circle cx="541" cy="388" r="7" fill="crimson"/><circle cx="541" cy="374" r="7" fill="crimson"/>
    <circle cx="541" cy="360" r="7" fill="crimson"/><circle cx="541" cy="346" r="7" fill="crimson"/>
  </g>

  <g class="fragment fade-in">
    <circle cx="451" cy="472" r="7" fill="crimson"/><circle cx="451" cy="458" r="7" fill="crimson"/>
    <circle cx="451" cy="444" r="7" fill="crimson"/><circle cx="451" cy="430" r="7" fill="crimson"/>
    <circle cx="451" cy="416" r="7" fill="crimson"/><circle cx="451" cy="402" r="7" fill="crimson"/>
    <circle cx="451" cy="388" r="7" fill="crimson"/>
    <circle cx="571" cy="472" r="7" fill="crimson"/><circle cx="571" cy="458" r="7" fill="crimson"/>
    <circle cx="571" cy="444" r="7" fill="crimson"/><circle cx="571" cy="430" r="7" fill="crimson"/>
    <circle cx="571" cy="416" r="7" fill="crimson"/><circle cx="571" cy="402" r="7" fill="crimson"/>
    <circle cx="571" cy="388" r="7" fill="crimson"/>
  </g>

  <g class="fragment fade-in">
    <circle cx="421" cy="472" r="7" fill="crimson"/><circle cx="421" cy="458" r="7" fill="crimson"/>
    <circle cx="421" cy="444" r="7" fill="crimson"/><circle cx="421" cy="430" r="7" fill="crimson"/>
    <circle cx="601" cy="472" r="7" fill="crimson"/><circle cx="601" cy="458" r="7" fill="crimson"/>
    <circle cx="601" cy="444" r="7" fill="crimson"/><circle cx="601" cy="430" r="7" fill="crimson"/>
  </g>

  <g class="fragment fade-in">
    <circle cx="391" cy="472" r="7" fill="crimson"/><circle cx="391" cy="458" r="7" fill="crimson"/>
    <circle cx="631" cy="472" r="7" fill="crimson"/><circle cx="631" cy="458" r="7" fill="crimson"/>
  </g>

  <g class="fragment fade-in">
    <circle cx="361" cy="472" r="7" fill="crimson"/>
    <circle cx="661" cy="472" r="7" fill="crimson"/>
  </g>

  <g class="fragment fade-in">
    <path d="M 340 478 C 420 478, 450 300, 511 300 S 602 478, 682 478"
          fill="none" stroke="#333" stroke-width="3"/>
    <text x="720" y="340" font-size="22" fill="#333" font-weight="bold">Approximately Normal</text>
  </g>
</svg>
```

## Takeaway

::: {.incremental}
- The **population** can look like anything. Here, it's skewed
- Each **individual sample mean** bounces around near the true mean
- The **distribution of sample means**, across many samples, is approximately $N\left(\mu, \dfrac{\sigma}{\sqrt{n}}\right)$
- Bigger $n$ → tighter, more normal-looking sampling distribution
:::


# Lets take it for a spin

<br>

> The total assets declared by general election candidates can be found on the ADR website. All the nearly 8000 candidates is actually the populaiton. We can decide on a sample size of 80 and see how well CLT works.

## Test CLT: an example

<br><br>

:::{.incremental}

1. Load data of all ~ 8000 general election candidates
2. draw a sample of size 80 (~1% of pop) and calcualte sample mean of assets.
3. repeat step 2 for about 1000 times
4. Plot density chart of all the 1000 sample means.
5. Does it look normal?
:::

## Test CLT: an example

**Load data**


```{r}
#| label: load-data-election
#| tbl-cap: Data from ADR on asset declared by general election candidates
#| output-location: slide
#| warning: false

read_csv("Lok-sabha-2024-candidates-details.csv") -> data_ele

data_ele |>
    select(sno, candidate, party, criminal_case, sum_all_movable) |>
    mutate(
        sum_all_movable = parse_number(sum_all_movable)
    ) |>
    reactable::reactable(defaultPageSize = 7) 
```