---
title: "Data, Evidence, and Decision Making (with R)"
author: "Ayush Patel"
subtitle: "Session 10: From Curiosity to evidence, a workflow"
institute: "National Law School of India University, Bengaluru"
date: today
date-format: "DD MMM, YYYY"
embed-resources: true
bibliography: citations.bib
execute:
  echo: true
  warning: false
format:
  revealjs:
    fig-height: 6
    fig-width: 14
    fig-dpi: 800
    embed-resources: true
    margin-left: 50px
    margin-right: 50px
    slide-number: c/t
    width: 1400
    height: 850
    theme: [default, custom.scss]
    footer: "Email: ayush.ap58@gmail.com"
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(janitor)

raw_names <- c("id","round_no","round_period","state","age","gender","income_bracket",
  "education","hh_size","earners","occupation",
  "econ_now","econ_outlook","employ_now","employ_outlook",
  "income_now","income_outlook","spend_now","spend_outlook",
  "essential_now","essential_outlook","nonessential_now","nonessential_outlook",
  "price_now","price_outlook","inflation_now","inflation_outlook",
  "inflation_rate_now","inflation_rate_now_pct","inflation_rate_future","inflation_rate_future_pct",
  "agri_land")

ccs <- read_csv("india_ccs_teaching.csv", skip = 1, col_names = raw_names, show_col_types = FALSE)

score3 <- function(x) case_when(
  x %in% c("Improved", "Will Improve", "Increased", "Will Increase") ~ 1,
  x %in% c("Worsened", "Will Worsen", "Decreased", "Will Decrease") ~ -1,
  x %in% c("Remained The Same", "Remain The Same") ~ 0,
  TRUE ~ NA_real_
)

order_lvls <- c("Decreased", "Remained The Same", "Increased")

ccs <- ccs |>
  mutate(
    essential_score = score3(essential_now),
    nonessential_score = score3(nonessential_now),
    income_score = score3(income_now),
    spend_score = score3(spend_now),
    essential_now = factor(essential_now, levels = order_lvls, ordered = TRUE),
    nonessential_now = factor(nonessential_now, levels = order_lvls, ordered = TRUE)
  )
```

## About Today's Dataset

The RBI's **Rural Consumer Confidence Survey**, May 2026 round. 8,813
households across 31 states and union territories. They were asked how
their economic situation, employment, spending, and prices have changed
compared to a year ago.

---

## A Note on Provenance

This is real RBI microdata. One respondent per row. It is not the
published index (CSI/FEI). RBI builds that index by aggregating responses
like these.

::: {.callout-note}
Everything you'll see today is the actual result on this data. That
includes whether a hypothesis holds up. Nothing here is staged.
:::

---

## Session 9 Recap

- Correlation is not causation
- A confounding variable can create an association that isn't real
- Simpson's Paradox: a trend can reverse once you split the data
- Randomisation and control groups exist to rule confounders out

---

## Today: From Abstract to Applied

Today those ideas stop being abstract. You'll use them to interrogate a
real dataset yourself. This isn't about learning five new R functions. It
is about practising a way of working.

---

## The Workflow

::: {.incremental}
1. **Understand the data**
2. **Ask a question**
3. **Form a hypothesis**
4. **Gather evidence**
5. **Interpret and refine**
:::

---

## The Workflow, Unpacked

::: {.incremental}
- What is a row, what is each variable?
- What relationship are you curious about?
- What do you expect, and why?
- Table it, chart it, quantify it, check it within groups
- Does the evidence hold up, or does it change your mind?
:::

We'll walk this loop once together on the survey data. Along the way,
you'll practise each step yourselves on a second, different dataset.

# Step 1: Understand the Data {background-color="#21295C"}

## Before Any Question: Look at the Data

You cannot ask a good question about data you haven't looked at.

- What does **one row** represent?
- Is each variable **categorical** or **continuous**?
- Are there units, ranges, or categories that surprise you?

---

## Look: Who Was Surveyed

```{r}
#| label: understand1
ccs |> select(state, age, gender, income_bracket, education, occupation) |> glimpse()
```

---

## Look: What They Were Asked

```{r}
#| label: understand2
ccs |> select(econ_now, econ_outlook, employ_now, employ_outlook, income_now, spend_now) |> glimpse()
```

---

## Look: `count()` on Occupation

```{r}
#| label: understand3
ccs |> count(occupation, sort = TRUE)
```

---

## Look: `count()` on Income Bracket

```{r}
#| label: understand4
ccs |> count(income_bracket, sort = TRUE)
```

---

## One More Step: Categories Aren't Numbers Yet

Every response here is a category. "Increased." "Decreased." "Remained
The Same." For a single number like a correlation, we still need a
numeric scale.

---

## From Categories to a Number

This is exactly how RBI itself builds its published indices. Net response
equals the share who said better, minus the share who said worse.

```{r}
#| label: recode
#| eval: false
score3 <- function(x) case_when(
  x %in% c("Increased", "Will Increase") ~  1,
  x %in% c("Decreased", "Will Decrease") ~ -1,
  x %in% c("Remained The Same", "Remain The Same") ~ 0
)
ccs <- ccs |> mutate(essential_score = score3(essential_now))
```

We'll use the categories directly for tables and charts, and this numeric
version only when we need a single summary number.

---

## Activity 1: Get to Know a New Dataset (5 min)

Switch datasets for this one. Use `diamonds`. It ships with `ggplot2`, so
it's already loaded. No import needed.

::: {.callout-tip title="Try it"}
1. Run `glimpse()` and `head()` on `diamonds`.
2. List three variables you find interesting, categorical or continuous.
3. What does a single row represent?
:::

Don't reach for `ggplot()` or `cor()` yet. This step is only about
orientation.

# Step 2: From Data to Questions {background-color="#065A82"}

## What Makes a Question Worth Asking?

A good exploratory question names **two things you suspect might be
related**. It is also answerable with the variables in front of you.

---

## Weak vs. Strong Questions

| Weak question | Stronger question |
|---|---|
| "What's in this survey?" | "Do households that report higher essential spending also report higher non-essential spending?" |
| "How do people feel about money?" | "Does that relationship look the same across income brackets?" |
| "Is there a pattern by state?" | "Does the economic outlook question track the employment outlook question?" |

---

## A Strong Question Names Its Columns

A strong question already tells you which columns you're about to look at
together. Often it also tells you which grouping variable might matter.

---

## Activity 2: Brainstorm and Pick a Question (5 min)

Stay with `diamonds` for this one.

::: {.callout-tip title="Try it"}
1. Brainstorm **three** questions of the form *"Does X relate to Y?"*
2. Pick the one you're most curious about.
3. Name the two columns it involves. Name any grouping variable too.
:::

---

## Our Shared Example

For the rest of today's instructor walkthrough we'll follow one example
on the survey data.

> *Does a household's change in essential spending track its change in
> non-essential spending?*

In the capstone, you'll run this exact loop again, start to finish, fully
on your own.

# Step 3: From Question to Hypothesis {background-color="#065A82"}

## Turning a Question Into a Hypothesis

A hypothesis, used loosely here, is your **specific, tentative guess**. It
has a direction and a reason. We'll formalise the term properly later in
the course.

---

## Our Hypothesis

> "I expect essential and non-essential spending to move together, because
> a household's overall spending capacity should affect both kinds of
> spending in the same direction."

---

## Just a Guess, For Now

::: {.callout-note}
Notice this is still just a guess. It might be wrong. It might be right
for the wrong reason. That's exactly what Step 4 is for.
:::

---

## Activity 3: Write Your Hypothesis (5 min)

Use your `diamonds` question from Activity 2.

::: {.callout-tip title="Try it"}
Write one sentence:

*"I expect [X] to [increase / decrease / relate to] [Y], because ___."*
:::

---

## Activity 3, Continued

Trade with a neighbour. Could their evidence gathering plan prove them
wrong?

# Step 4: Gather Evidence {background-color="#1C7293"}

## Evidence, Instrument 1: Build a Two-Way Table

Both variables here are categories, not numbers. The natural first move is
a table, not a scatterplot. A two-way table counts every combination of
responses.

---

## Counts and Proportions, Together

The `janitor` package builds this kind of table directly. Each cell shows
a percentage of the grand total, with the raw count right beside it. Row
and column totals come along too.

```{r}
#| label: evidence-table
ccs |>
  tabyl(essential_now, nonessential_now) |>
  adorn_totals(c("row", "col")) |>
  adorn_percentages("all") |>
  adorn_pct_formatting(digits = 1) |>
  adorn_ns()
```

---

## Evidence, Instrument 2: Chart the Proportions

A table of numbers is precise but slow to read. Two kinds of bar chart
make the same numbers easier to see: one for absolute counts, one for
relative shares.

---

## Absolute Counts

```{r}
#| label: evidence-bar-absolute
ggplot(ccs, aes(x = essential_now, fill = nonessential_now)) +
  geom_bar(position = "dodge") +
  labs(x = "Essential spending, vs. year ago",
       y = "Number of households",
       fill = "Non-essential spending")
```

---

## Relative Proportions

```{r}
#| label: evidence-bar-relative
ggplot(ccs, aes(x = essential_now, fill = nonessential_now)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Essential spending, vs. year ago",
       y = "Share of households",
       fill = "Non-essential spending")
```

---

## Read the Pattern

Households who say essential spending increased are also more likely to
say non-essential spending increased. Is that a real relationship, or
just two variables that both trend upward on their own?

---

## Evidence, Instrument 3: Quantify It

A table and a chart both give you a pattern. A correlation coefficient
turns that pattern into one number, using the numeric scores from Step 1.
It is useful, but it is only ever a summary.

---

## Picking the Right Correlation

Our scores are ordinal. "Increased" is more than "Remained The Same,"
which is more than "Decreased." But we can't say the gap between the
first two equals the gap between the last two.

---

## Why That Rules Out Pearson

Pearson's correlation, the default in `cor()`, assumes those gaps are
equal and meaningful. That assumption isn't ours to make here.

Spearman's correlation only uses the ranking of the categories. It never
needs the size of the gap between them. That makes it the safer choice for
ordinal data like this.

---

## Compute the Correlation

```{r}
#| label: evidence-cor
cor(ccs$essential_score, ccs$nonessential_score, method = "spearman", use = "complete.obs")
```

---

## A Weak Positive. Now What?

::: {.callout-important}
Weakly positive, pooled across all 8,813 households. Before deciding what
that means, get suspicious of a single pooled number.
:::

## Classroom Activity: Spot the Confounder (15 min)

Before you trust that correlation number, get suspicious of it.

---

## Three Correlations

::: {.callout-tip title="In small groups: find a confounder for each"}
1. Ice cream sales correlate with drowning incidents.
2. Firefighters at a fire correlate with total damage cost.
3. Shoe size correlates with reading ability, in children.
:::

---

## Then Discuss

**What data would you need to collect to rule out that confounder?**

This is the exact scepticism you're about to apply to our spending
result.

---

## Evidence, Instrument 4: Check It Within Groups

Households across income brackets face very different spending
constraints. Does `income_bracket` change the essential and non-essential
story?

---

## Proportions, Faceted by Income Bracket

```{r}
#| label: evidence-group-bar
ggplot(ccs, aes(x = essential_now, fill = nonessential_now)) +
  geom_bar(position = "fill") +
  facet_wrap(~ income_bracket, nrow = 2) +
  scale_y_continuous(labels = scales::percent) +
  theme(legend.position = "bottom") +
  labs(x = "Essential spending, vs. year ago",
       y = "Share of households",
       fill = "Non-essential spending")
```

---

## Correlation, Within Each Income Bracket

```{r}
#| label: evidence-group-cor
ccs |>
  group_by(income_bracket) |>
  summarise(n = n(),
            r = cor(essential_score, nonessential_score,
                    method = "spearman", use = "complete.obs")) |>
  arrange(r)
```

## What Just Happened? The Numbers

- **Pooled correlation:** `r round(cor(ccs$essential_score, ccs$nonessential_score, method="spearman", use="complete.obs"), 2)`
- **Within income brackets:** ranges from **0.12** to **0.41**
- Weakest in the middle brackets. Strongest at both extremes.

---

## What Just Happened? Not a Reversal, But Still a Story

::: {.incremental}
- This isn't a sign flip like classic Simpson's Paradox. The relationship
  never goes negative.
- The pooled number of 0.16 still hides something real. The strength of
  the relationship depends heavily on income bracket.
- The poorest households may cut or grow both spending types together out
  of necessity. The richest may do the same out of shared lifestyle
  shifts.
- The middle brackets have more room to trade one off against the other.
  That's a plausible read, not a proven one.
:::

---

## What Just Happened? The Lesson

::: {.incremental}
- Confounding doesn't always look like a dramatic reversal.
- Sometimes a hidden grouping variable changes *how strong* a relationship
  is, not just its direction.
- Either way, the fix is the same. Never trust a single pooled number.
:::

# Step 5: Interpret and Refine {background-color="#21295C"}

## Closing the Loop: What We Learned

::: {.incremental}
- The evidence didn't just confirm or reject the hypothesis. It
  **sharpened** it. Essential and non-essential spending move together
  weakly overall, and more strongly at the income extremes.
- Income bracket matters for how tightly these two move together.
:::

---

## Closing the Loop: What's Still Missing

::: {.incremental}
- We still can't say one causes the other. For that you'd need Session
  9's tools. Randomisation, control groups, ruling out other confounders.
- We haven't ruled out other explanations, like state, occupation, or
  household size. Checking one grouping variable doesn't rule out the
  rest.
- Every hypothesis should end with a question. *What would I still need
  to check?*
:::

## Readings

- R4DS 2e, Ch. 10, *Exploratory Data Analysis*
- IMS 2e, Ch. 7, *Linear Regression*

Read these as reference for the instruments. The workflow around them is
what today's session was about.

# Final Authentic Task {background-color="#21295C"}

## Capstone: Go Exploring, On Your Own Terms

Same dataset, `ccs`. But **don't reuse essential and non-essential
spending, and don't reuse `income_bracket`.** Everything else is open.

---

## The Task: Steps 1 to 3

::: {.callout-important title="Steps 1 to 3"}
1. **Understand it.** Look again, with fresh eyes, at columns you haven't
   touched. There are over 30.
2. **Ask your own question.** Try any two of the perception and outlook
   columns, or one against `state`, `occupation`, `gender`, `age`, or
   `education`.
3. **Write a hypothesis.** One sentence, a direction, and a reason.
:::

---

## The Task: Steps 4 and 5

::: {.callout-important title="Steps 4 and 5"}
4. **Gather evidence.** Recode your variables to a numeric score if
   needed. Then build a two-way table, a bar chart of proportions, a
   pooled correlation, and a within-group correlation against at least
   one grouping variable. Use Spearman's method. These are ordinal
   variables.
5. **Interpret.** Does the evidence support, sharpen, or overturn your
   hypothesis? What would you still need to claim causation?
:::

---

## No Answer Key

There is no single "correct" pattern hiding in this dataset for you to
find. Some variable pairs will show a real relationship. Some won't. Some
groupings will change the story. Some won't. Report what you actually
find.

## Wrap-Up: The Workflow

**Understand, then Question, then Hypothesise, then Gather Evidence, then
Interpret.**

The table, the chart, and the correlation are instruments. They only
matter once you know what you're testing.

---

## Wrap-Up: The Lesson

- A pooled number can hide a confounder. Sometimes that's a reversal.
  Sometimes it's just a change in strength.
- Real data doesn't owe you a dramatic result. A weak or inconsistent
  finding is still a finding.
- Every round of evidence should leave you asking what you'd still need to
  check.

**Next session:** formalising this with linear regression models.
