---
title: "Numbers"
author: "Ayush Patel"
subtitle: Session 6
institute: "National Law School of India University, Bengaluru"
date: today
date-format: "DD MMM, YYYY"
embed-resources: false
bibliography: citations.bib
execute: 
  echo: false
  warning: false
format: 
  revealjs:
    fig-height: 6
    fig-width: 14
    fig-retina: 1
    embed-resources: true
    margin-left: 50px
    margin-right: 50px
    slide-number: c/t
    width: 1400
    height: 850
    theme: [default, custom.scss]
    footer: "Email: ayush.ap58@gmail.com"
---

```{r}
#| include: false
#| warning: false

extrafont::loadfonts(device = "win")

library(tidyverse)
library(hrbrthemes)
```

# Random Process

<br>

> An event for which all possible outcomes are known but it is uncertain which particular outcome will occur.

# Probability

<br>

> The proportion of occurrences of an outcome of a random process observed over a large number of observations.

<br>
Why large number of observation?

# Probability Distribution

<br>

1. The sample distribution is not an accurate picture of population distribution

2. We need a way to arrive at a reasonable estimate of the probability of outcomes in population using sample data

3. If the sample is representative, we can estimate the probability distribution of a random process. This distribution is called the probability distribution.

# Random Variable

<br>

> The outcomes of a Random Process can be assigned with a value. Recordings of the values of the outcomes of a random process is called a Random Variable

<br><br>

There can be discrete or continuous random variables.

## Probability Distribution of discrete random variables

<br>

For an outcome space of a random process A, defined as $A  = \{a| a \space is \space number \space of \space instagram followers\}$

<br>

$0 <= P(a) <= 1$ and $\sum_{i =1}^{|A|}{P(a_i)} = 1$

## Car Ownership

A survey is given to all of section E students. With only one question. Number of cars owned by their HH. Following is the summary table:

|Number of Cars owned (N) | Total HH that own N cars | % of HH that own N cars |
|-------------------------|--------------------------|-------------------------|
| 0 | 4 | 5.7% |
| 1 | 27 | 38.6% |
| 2 | 18 | 25.7% |
| 3 | 15 | 21.4% |
| 4 | 6 | 8.6 % |

## Car Ownership

```{r}
tibble(
  n = c(0:4),
  resp = c(4,27,18,15,6),
  prop = resp/70 
) |>
  ggplot(aes(n,prop)) +
  geom_col(width = 1, 
  colour = "black",
  fill = "lightgrey") +
  labs(
    x = "Number of cars owned by HH",
    y = "Probablity",
    title = "Probablity Distribution of cars owned by HH"
  )+
  theme_ipsum_es()
```

## Probability Distribution of a continious random variable

<br>

1. The outcome space of a continious random variable can be infinite. Ex: Weight of all elephants. Note the range may be finte, but the outcome space can be infinite.

2. So, we must use intervals of a continious random variable to estimate probablity at a point. Say you want probability of the value 3420. So we estiate the probablity of the interval 3420 - a very tiny number to 3420 + a very tiny number. This is done using calculating the area under the cure.  

# An intuitive tour of distributions

<br>

> [Seeing Theory Chapter 3](https://seeing-theory.brown.edu/#secondPage/chapter3) without the central limit theorm

# Review Questions

1. Agresti Ch4, question 4.3
2. Agresti Ch4, question 4.4
3. Introductory Stats Ch4, 4.4tryit
4. Introductory Stats Ch4, 4.5tryit5

# More Normal Stuff

> The normal distribution needs more attention because we find many naturally occurring numbers take this shape and because it is very useful for statistical inference.